---
title: "Grays Harbor Recreational Crab Harvest Estimate"
subtitle: "Pooled CPUE Model - 2024-25 Season Results Report"
author: "Washington Department of Fish and Wildlife"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: false
    code_folding: hide
    theme: cosmo
    highlight: tango
knit: >
  (function(input, ...) {
    d <- format(Sys.Date(), "%Y%m%d")
    outdir <- file.path(dirname(dirname(normalizePath(input))), "05_output", d, "pooled-CPUE")
    dir.create(outdir, recursive = TRUE, showWarnings = FALSE)
    rmarkdown::render(input, output_dir = outdir, envir = new.env())
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", fig.width = 10, fig.height = 6)
# A small helper to render a data frame as a clean, captioned report table.
report_table <- function(df, caption = NULL, digits = 2) {
  if (is.null(df) || !is.data.frame(df) || nrow(df) == 0) return(invisible(NULL))
  knitr::kable(df, caption = caption, digits = digits, format.args = list(big.mark = ","))
}

# Short, HTML-friendly fit labels for display tables (e.g. "boat all-gear").
short_fit <- function(x) gsub("_Dungeness_Kept$", "",
                         gsub("private_boat", "boat",
                           gsub("_ring_net_only", " pot-closure",
                             gsub("_all_gear", " all-gear", x))))

# Pot-closure boundary markers for date-axis plots. Draws a dotted line at the pot-
# closure start and a dashed line at the day pots open (closure end + 1), so every
# season plot brackets the pot-closure window explicitly (added 2026-07-13 with the
# explicit pot_closure_start/end config). Returns a list of ggplot layers, added to a
# plot with `+`. When the closure starts at the season start the left line sits at the
# left edge (correct). labels = FALSE draws the lines only (for secondary panels).
pot_closure_markers <- function(params, labels = TRUE) {
  pc_start  <- as.Date(params$pot_closure_start %||% params$est_date_start)
  pots_open <- as.Date(params$pot_open_date     %||% (as.Date(params$pot_closure_end) + 1))
  layers <- list(
    geom_vline(xintercept = pc_start,  linetype = "dotted", color = "darkgreen", linewidth = 0.5),
    geom_vline(xintercept = pots_open, linetype = "dashed", color = "red",       linewidth = 0.5))
  if (labels) layers <- c(layers, list(
    annotate("text", x = pc_start + 3,  y = Inf, label = "Pot closure start",
             hjust = 0, vjust = 1.5, size = 3, color = "darkgreen"),
    annotate("text", x = pots_open + 3, y = Inf, label = "Pots open",
             hjust = 0, vjust = 1.5, size = 3, color = "red")))
  layers
}

# Read the per-fit CSVs a section wrote to output_dir (files named prefix_<label>.csv)
# and row-bind them for on-page display, tagging each row with a short fit_lbl label from
# the filename. Display-only and defensive: returns NULL if output_dir or the files are
# absent (a PE-only fit writes none), so a display chunk can never break the knit.
read_fit_outputs <- function(prefix, dir = NULL) {
  if (is.null(dir)) {
    if (!exists("output_dir", inherits = TRUE)) return(NULL)
    dir <- get("output_dir", inherits = TRUE)
  }
  fs <- list.files(dir, pattern = paste0("^", prefix, "_.*\\.csv$"), full.names = TRUE)
  if (!length(fs)) return(NULL)
  tryCatch(
    dplyr::bind_rows(lapply(fs, function(f) {
      d <- suppressWarnings(utils::read.csv(f, stringsAsFactors = FALSE, check.names = FALSE))
      d$fit_lbl <- short_fit(sub(paste0("^", prefix, "_"), "",
                              tools::file_path_sans_ext(basename(f))))
      d
    })),
    error = function(e) NULL)
}
```

> **How to read this report.** This is the results report for one seasonal harvest estimate from the pooled CPUE model. Code is hidden by default; use the **Code** buttons (or *Code > Show All Code* at the top right) to reveal any step. Each section states what it does, shows the notable tables and figures, and, for the diagnostic sections, notes what a healthy run should look like. The headline number is the port total in the Results section; always read it together with the convergence and PE-vs-BSS sections, which say how much to trust it.


## 1. Setup and configuration

Loads packages, sources the helper functions in `03_R_functions/`, applies the run configuration from `run_config.R`, and defines the two sub-seasons (pot-closure and all-gear). Every user-selectable toggle for this run lives in `run_config.R`.

```{r config, results='hide'}
###############################################################################
# Crab Creel Estimation - Westport / Grays Harbor  |  Pooled CPUE Model
# 2024-25 Season  |  Pipeline code v7.9 (2026-07-12)
#
# Estimates total recreational Dungeness crab harvest at Westport / Grays Harbor
# from a single pooled CPUE process shared across gear types. Three population
# components are estimated independently and summed to the port total:
#   1. Shore crabbers (dock + jetty + beach)   BSS + PE
#   2. Private boat crabbers                    BSS + PE
#   3. Commercial/charter boats                 census/tally
# Two gear-regime sub-seasons: pot closure (Sep 16 - Nov 30, non-pot gear only) and
# all-gear (Dec 1 - Sep 15, pots allowed). "Pot closure" was formerly labeled
# "ring-net only"; non-pot gear other than ring nets is also legal, so the name was
# a misnomer. The internal key stays "ring_net_only" for output-filename continuity.
#
# CURRENT STATE (v7.9): both fitted components run on the gear-DEPLOYMENT effort
# unit (h = number_of_gear, L = tau), via the shared 03_R_functions/bss_effort_spec.R.
# Helper functions live in 03_R_functions/ (sourced wholesale in the setup chunk),
# and every user-selectable toggle lives in run_config.R (the single file you edit);
# this .Rmd's params block holds only model-internal tuning (sampler settings, gate
# and AR-selector thresholds).
#
# KEY FEATURES: adaptive AR(1) temporal resolution; B1_C weekend CPUE effect;
# L_effective day length estimated with uncertainty from I/E; data-driven R_G prior;
# gear-deployment effort scale (R_G_boat / tau_boat, tau_shore); scale-aware
# convergence gate; expected + predictive catch in the generated quantities.
#
# FULL CHANGE LOG (every version, newest first, with working notes):
#   07_documentation/BSS-GH-pooled-CPUE-model-development-history.md
# METHOD REFERENCE (the published method of record):
#   07_documentation/BSS-GH-pooled-CPUE-model-documentation.md
###############################################################################

# ===========================================================================
# 0. SETUP
# ===========================================================================

load.lib <- c("tidyverse","lubridate","suncalc","gt","patchwork","rstan","here","readxl")
install.lib <- load.lib[!load.lib %in% installed.packages()]
for(lib in install.lib) install.packages(lib, dependencies=TRUE)
sapply(load.lib, require, character=TRUE)
rstan_options(auto_write = TRUE)
purrr::walk(list.files(here("03_R_functions"), full.names = TRUE), source)

run_date <- format(Sys.Date(), "%Y%m%d")
output_dir <- here("05_output", run_date, "pooled-CPUE")
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

# timer_start() / timer_stop() and their timing state now live in
# 03_R_functions/bss_timers.R (auto-sourced). Read collected timings with
# bss_timer_log(). Extracted; behavior unchanged.

run_start_time <- Sys.time()

# ===========================================================================
# 0.1 PARAMETERS
# ===========================================================================

params <- list(
  # NOTE: every user-selectable toggle (season window, structural dates, catch
  # groups, day-typing, effort unit, incomplete-trip filter, I/E settings,
  # holidays, and the model-behavior levers such as collapse_mu_hier) now lives in
  # run_config.R, the single source of truth, and is applied as an override below.
  # This list holds ONLY this model's internal tuning: the Stan model file, the
  # per-fit sampler settings, the convergence-gate thresholds, and the adaptive
  # AR-selector thresholds. These rarely change and legitimately differ from the
  # gear-resolved model, so they stay with the model rather than in run_config.

  # --- Stan model file ---
  bss_model_file    = "crab_bss_pooled.stan",

  # --- BSS sampler defaults ---
  bss_adapt_delta   = 0.9,

  # --- BSS convergence tuning (extra iterations for longer sub-seasons / daily AR) ---
  bss_iter_default          = 3000,
  bss_warmup_default        = 1500,
  bss_max_treedepth_default = 10,
  bss_iter_shore_allgear    = 5000,
  bss_warmup_shore_allgear  = 2500,

  # Per-fit treedepth / adapt_delta overrides (v6.1, ported from gear-resolved
  # v5.2). The shore all-gear daily-AR posterior is the densest geometry and
  # saturated max_treedepth=10 on 91.6% of iterations in the 2026-04-08 run;
  # ring-net fits are data-sparse and benefit from tighter adaptation. Any fit
  # without an override uses the *_default values.
  bss_treedepth_shore_allgear = 14,
  bss_delta_shore_allgear     = 0.95,
  bss_treedepth_ringnet       = 12,
  bss_delta_ringnet           = 0.95,

  # v6.2: boat all-gear sampler tuning. The boat all-gear fit failed in the
  # 2026-04-08 run (5998 divergences, R-hat 1.074, n_eff 76) under defaults.
  # treedepth_pct was 0, so divergences came from step size: adapt_delta is
  # raised to 0.99 (primary lever) and max_treedepth to 13 (buffer), with more
  # iterations to give n_eff a chance to clear 400.
  bss_iter_boat_allgear       = 5000,
  bss_warmup_boat_allgear     = 2500,
  bss_treedepth_boat_allgear  = 13,
  bss_delta_boat_allgear      = 0.99,

  bss_min_interviews = 20,

  # --- B1.8: SCALE-AWARE convergence gate thresholds --------------------------
  # The gate rejects a BSS fit whose divergent transitions have BIASED the
  # reported totals, while accepting a fit whose divergences are diffuse and
  # harmless (Betancourt 2017). It measures the divergent draws' impact on each
  # reported TOTAL in units of that total's posterior SD (scale-invariant):
  #   impact = |median(all_draws) - median(bulk_draws)| / sd(all_draws)
  # A fit passes when max(impact_C, impact_E) is below max_divergence_total_impact_sd,
  # with a hard divergent-FRACTION backstop (max_divergence_fraction). See the
  # model documentation and development history for the full B1.8 rationale.
  max_divergence_total_impact_sd = 0.10,
  # P7 (2026-07-12): tightened from 0.15 to 0.05. The scale-aware impact test above
  # (max_divergence_total_impact_sd) is the primary gate; this fraction is only a hard
  # backstop, and 0.15 was too loose to defend for a publication method (a fit at ~14%
  # divergences would have passed as BSS). 0.05 matches the gear-resolved model. The
  # 2026-07-11 pooled fits (0.54%, 2.23%, 2.5% divergences) are all under 5%, so this
  # changes no current gating decision; divergences indicate biased HMC exploration
  # (Betancourt 2017), so the backstop belongs near 5%.
  max_divergence_fraction        = 0.05,

  # --- DEPRECATED gate parameters (retained for reference; no longer gating) ---
  # Superseded by the scale-aware gate above. The level-normalized distortion is
  # still COMPUTED and reported in convergence_report.csv (distortion_C /
  # distortion_E) for continuity, but neither value participates in pass/fail.
  max_divergences = 5,
  max_divergence_distortion = 0.02,

  # --- Adaptive AR temporal-resolution thresholds -----------------------------
  # The model selects the finest AR resolution the data can support:
  #   daily:   effort obs on >= ar_daily_min_coverage of days AND >= ar_daily_min_obs
  #   weekly:  >= ar_weekly_min_per_week effort obs per week on average
  #   monthly: fallback for sparse data
  ar_daily_min_coverage  = 0.25,
  ar_daily_min_obs       = 20,
  ar_weekly_min_per_week = 1.5,

  # v6.5 (B1): per-population cap on AR resolution. The map now lives in run_config.R
  # (ar_max_resolution$pooled) so both models' caps sit in one place; this model's cap
  # is selected from it just after the merge below (P5, 2026-07-12). The boat
  # (trailer-count) effort series is too weakly informative for daily AR, so the boat
  # is capped at weekly; a population absent from the map defaults to "daily" (no cap).

  # --- I/E diagnostics ---
  ie_use_all_dates_for_diagnostics = TRUE
)

# --- Run-level config from run_config.R (single source of truth) --------------
# run_config.R owns every user-selectable toggle (season window, structural dates,
# catch groups, day-typing, effort unit, filters, I/E settings, holidays, the AR
# resolution map, and the model-behavior toggles). When launched via run_estimation.R
# the orchestrator has already defined `run_config`; when this .Rmd is knit standalone
# the guard below sources run_config.R so the same toggles are used either way.
if (!exists("run_config")) source(here::here("run_config.R"))
# P5 (2026-07-12): run_config is the BASE parameter set; this model's internal tuning
# (defined above) is layered on top. The two key sets are disjoint, so the merge order
# carries intent only (run_config is the base, the model updates it with its specifics).
params <- modifyList(run_config, params)
# Select this model's AR-resolution cap from the shared per-model map in run_config.
params$ar_max_resolution <- run_config$ar_max_resolution$pooled

# Crabbing holidays come from run_config (single source of truth). Kept as a bare
# object so any inline reference in this driver still resolves.
crabbing_holiday_dates <- params$crabbing_holiday_dates


catch_groups <- if(params$estimate_red_rock) {
  c("Dungeness_Kept", "Red_Rock_Kept")
} else {
  "Dungeness_Kept"
}

# Sub-seasons split by the pot-closure window (03_R_functions/build_subseasons.R).
# The pot closure is when pots are not legal (non-pot gear only); its window is set
# explicitly in run_config (pot_closure_start / pot_closure_end). For the current
# config (closure at the season start) this returns the historical two sub-seasons
# with unchanged internal names ("ring_net_only" is the pot-closure key, kept for
# output-filename continuity; display_name is "Pot closure"), so keys and filenames
# do not change. A mid-season closure adds pre/post all-gear periods automatically.
# Model logic keys off ss$gear_regime ("pot_closure" | "all_gear"), never ss$name.
subseasons <- build_subseasons(params)
for (ss in subseasons) {
  cat(sprintf("Sub-season: %-26s %s to %s (%d days)\n",
              paste0(ss$display_name, " [", ss$name, "]"),
              ss$start, ss$end, as.integer(ss$end - ss$start + 1)))
}
cat(sprintf("Catch groups: %s\n", paste(catch_groups, collapse = ", ")))

# Sub-season aggregation is now GENERAL (2026-07-13): the PE port summary, the catch-by-
# mode table, and the final harvest table sum over whatever sub-seasons build_subseasons
# produced (via pe_pop_sum / pe_port_total below), rather than the hardcoded historical
# pair. A mid-season pot closure (which adds all_gear_pre / all_gear_post keys) is
# therefore handled end-to-end. That general path has NOT yet been validated against a
# fitted run, so emit a loud note when the sub-season set departs from the historical
# pair; verify the port total against the per-component PE table before citing such a run.
ss_name_set <- vapply(subseasons, function(z) z$name, character(1))
if (!setequal(ss_name_set, c("ring_net_only", "all_gear"))) {
  cat(sprintf(paste0(
    "\n*** NOTE: non-historical sub-season set {%s}. The report aggregation sums over all ",
    "sub-seasons per population, but this configuration has not been validated against a ",
    "fitted run; verify the port total against the per-component PE table before citing. ***\n"),
    paste(ss_name_set, collapse = ", ")))
}

# --- Generalized PE aggregation over the sub-season set (build_subseasons) -----------
# Sum a PE field over all sub-seasons of a population so the port total, the by-mode
# table, and the final-table PE rows follow build_subseasons instead of a fixed 5-row
# layout. These close over the run globals `pe_all` (populated by the PE loop) and
# `subseasons`; they are only called after the PE loop, so pe_all exists by call time.
# pe_all keys are "<population>_<subseason-name>"; a missing key or field contributes 0.
pe_field   <- function(key, field) pe_all[[key]][[field]] %||% 0
pe_pop_sum <- function(pop, field)
  sum(vapply(subseasons, function(ss) pe_field(paste0(pop, "_", ss$name), field), numeric(1)), na.rm = TRUE)
pe_port_total <- function(field)  # na.rm matches the baseline sum([1:5], na.rm=TRUE):
  sum(pe_pop_sum("shore", field), pe_pop_sum("private_boat", field),
      pe_field("comm_charter", field), na.rm = TRUE)


# ===========================================================================
# 0.2 STANDALONE DAY-TYPE CLASSIFIER
#
# Assigns day_type to ANY date, independent of the estimation calendar.
# Use this for diagnostic plots that may show data outside the estimation
# period (e.g., effort data collected year-round but est_date_end is Jan 31).
# ===========================================================================

# classify_day_type() now lives in 03_R_functions/classify_day_type.R
# (auto-sourced). Signature is now (dates, params); it reads weekends/holidays
# from params. Behavior unchanged.


# ===========================================================================
# 1. DATA INGESTION
# ===========================================================================

# fetch_crab_data() now lives in 03_R_functions/fetch_crab_data.R
# (auto-sourced). Extracted verbatim; behavior unchanged.


# ===========================================================================
# 1.5 INGRESS/EGRESS DATA
# ===========================================================================

# fetch_ie_data() now lives in 03_R_functions/bss_day_length.R, shared with the
# gear-resolved driver so the two cannot drift. It is auto-sourced by the
# purrr::walk over 03_R_functions/ near the top of this file. Function body
# moved verbatim; behavior unchanged.


# ===========================================================================
# 1.6 L_EFFECTIVE MODEL - Regression with uncertainty propagation
#
# Fits a
# regression of log(L_effective) on day-of-year (quadratic) and day type.
# This captures the seasonal gradient WITHIN sub-seasons and provides
# per-day prediction uncertainty for propagation into the Stan model.
#
# Returns: list with
#   $predict_fn: function(event_date, day_type) -> tibble(L_mu, L_sigma)
#   $model: the fitted lm object
#   $detail: per-I/E-day data
#   $summary: model coefficients and fit statistics
# ===========================================================================

# estimate_L_effective() now lives in 03_R_functions/bss_day_length.R, shared
# with the gear-resolved driver so the two cannot drift. It is auto-sourced by
# the purrr::walk over 03_R_functions/ near the top of this file. Function body
# moved verbatim; behavior unchanged.


# ===========================================================================
# 2. PREP DAYS - Build calendar with L_mu and L_sigma per day
#
# Stores both L_mu (predicted median) and L_prior_sigma
# lookup value, store both L_mu (predicted median) and L_prior_sigma
# (prediction uncertainty on log scale) for each day. These feed into
# the Stan model where L becomes a parameter.
# ===========================================================================

# prep_days_crab() now lives in 03_R_functions/prep_days_crab.R (shared,
# auto-sourced). Signature is now (date_begin, date_end, params, L_eff_model);
# it derives weekends/holidays/period_pe/sections from params. Behavior unchanged.


# ===========================================================================
# 3. SUMMARIZE DATA - Filter to population x sub-season
# ===========================================================================

# prep_population_summary() now lives in
# 03_R_functions/prep_population_summary.R (shared, auto-sourced). catch_groups is
# derived from params inside. Behavior unchanged.


# ===========================================================================
# 4. PE METHOD - (unchanged from v1)
# ===========================================================================

# run_pe_pooled() now lives in 03_R_functions/run_pe_pooled.R (auto-sourced).
# catch_groups is derived from params; the shore branch now flows through
# bss_effort_spec() so shore PE matches the shore BSS unit (2026-07-11 fix).


# ===========================================================================
# 5. BSS PREP - Build Stan data list for crab_bss_pooled.stan
#
# ADAPTIVE AR RESOLUTION: Selects daily, weekly, or monthly AR based on
# effort data density. Dense data → daily AR for smooth interpolation.
# Sparse data → coarser resolution to avoid poorly-identified AR dynamics.
#
# Also includes: data-driven R_G prior, L_effective as parameter (shore),
# informative R_G_boat prior, tightened Cauchy scales, no interview cap.
# ===========================================================================

# prep_bss_crab_pooled() now lives in 03_R_functions/prep_bss_crab_pooled.R
# (auto-sourced). Extracted verbatim (renamed); behavior unchanged.


# ===========================================================================
# 6. COMMERCIAL/CHARTER CENSUS (unchanged from v1)
# ===========================================================================

# estimate_comm_charter() now lives in 03_R_functions/estimate_comm_charter.R
# (shared, auto-sourced). crabbing_holiday_dates read from params. Behavior unchanged.


# ===========================================================================
# 7. RUN EVERYTHING
# ===========================================================================

```

## 2. Data ingestion

Reads the season's effort counts, crabber interviews, commercial tally, and ingress/egress (I/E) surveys, and classifies each interview into shore, private-boat, or commercial/charter.

```{r load-data, results='hide'}
# ---- 7.0 LOAD DATA ----
timer_start("1_load_data")
cat("\n", strrep("=",60), "\n LOADING DATA\n", strrep("=",60), "\n")
dwg <- fetch_crab_data(params)
ie_data <- fetch_ie_data(params)

# --- Fit L_effective regression ---
L_eff_model <- NULL
if(isTRUE(params$use_ie_day_length) && nrow(ie_data) > 0) {
  L_eff_model <- estimate_L_effective(ie_data, params$pot_open_date, params)
  if(!is.null(L_eff_model)) {
    # Save regression diagnostics
    if(!is.null(L_eff_model$detail)) {
      write.csv(L_eff_model$detail,
                file.path(output_dir, "L_effective_ie_detail.csv"), row.names = FALSE)
    }
  }
}
timer_stop("1_load_data")


```

## 3. Input diagnostics: effort, CPUE, and day length

Diagnostic views of the raw inputs before any modeling: sampled effort over the season by day type, interview CPUE over time, and the effective day length derived from the I/E surveys. Use these to spot data gaps or anomalies (for example a run of unsampled weekends, or an implausible CPUE spike).

```{r input-figures, results='hide', fig.show='hide'}
# ---- 7.1 DIAGNOSTIC PLOTS ----
timer_start("2_diagnostic_plots")
cat("\n", strrep("=",60), "\n DIAGNOSTIC PLOTS\n", strrep("=",60), "\n")

days_full <- prep_days_crab(params$est_date_start, params$est_date_end,
                            params, L_eff_model = L_eff_model)

# Plot 1: Shore gear count timeseries
p_shore_effort <- dwg$shore_effort |>
  group_by(event_date) |>
  summarise(total_gear = sum(count_quantity), n_counts = n(), .groups = "drop") |>
  mutate(day_type = classify_day_type(event_date, params)) |>
  ggplot(aes(x = event_date, y = total_gear, color = day_type, size = n_counts)) +
  geom_point(alpha = 0.5) +
  pot_closure_markers(params, labels = TRUE) +
  scale_size_continuous(range = c(1,3), breaks = 1:3, name = "Counts/day") +
  labs(title = "Shore: Daily Gear Count (Docks F20 + F17-21)",
       x = "Date", y = "Total gear count", color = "Day type") +
  theme_bw()
ggsave(file.path(output_dir, "plot_shore_effort_timeseries.png"), p_shore_effort, width = 10, height = 6)

# Plot 2: Boat trailer count timeseries
p_boat_effort <- dwg$boat_effort |>
  group_by(event_date) |>
  summarise(total_trailers = sum(count_quantity), n_counts = n(), .groups = "drop") |>
  mutate(day_type = classify_day_type(event_date, params)) |>
  ggplot(aes(x = event_date, y = total_trailers, color = day_type)) +
  geom_point(alpha = 0.5) +
  pot_closure_markers(params, labels = FALSE) +
  labs(title = "Private Boat: Daily Trailer Count", x = "Date", y = "Trailers", color = "Day type") +
  theme_bw()
ggsave(file.path(output_dir, "plot_boat_effort_timeseries.png"), p_boat_effort, width = 10, height = 6)

# Plot 3: CPUE by population
cpue_data <- dwg$interview |>
  filter(fishing_time_total > 0, dungeness_kept >= 0) |>
  mutate(cpue = dungeness_kept / fishing_time_total) |>
  group_by(event_date, population) |>
  summarise(mean_cpue = mean(cpue), n_int = n(), .groups = "drop") |>
  mutate(day_type = classify_day_type(event_date, params)) |>
  filter(population %in% c("shore", "private_boat"))

p_cpue <- cpue_data |>
  ggplot(aes(x = event_date, y = mean_cpue, color = day_type)) +
  geom_point(alpha = 0.4, size = 1.5) +
  geom_smooth(method = "loess", span = 0.3, se = TRUE, color = "grey30") +
  pot_closure_markers(params, labels = FALSE) +
  facet_wrap(~population, ncol = 1, scales = "free_y") +
  labs(title = "Dungeness CPUE by Population", x = "Date", y = "Mean CPUE", color = "Day type") +
  theme_bw()
ggsave(file.path(output_dir, "plot_cpue_timeseries.png"), p_cpue, width = 10, height = 8)

# Plot 4: Gear counts by month and day type (boxplot)
p_effort_month <- dwg$shore_effort |>
  group_by(event_date) |>
  summarise(total_gear = sum(count_quantity), .groups = "drop") |>
  mutate(day_type = classify_day_type(event_date, params),
         month_label = factor(format(event_date, "%Y-%m"))) |>
  ggplot(aes(x = month_label, y = total_gear, fill = day_type)) +
  geom_boxplot() +
  labs(title = "Shore Gear Counts by Month and Day Type",
       x = "Month", y = "Daily gear count", fill = "Day type") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
ggsave(file.path(output_dir, "plot_effort_by_month.png"), p_effort_month, width = 12, height = 6)

# L_effective regression diagnostic plot
if(!is.null(L_eff_model) && !is.null(L_eff_model$detail) && nrow(L_eff_model$detail) >= 3) {
  # Predict L for a full year of days for visualization
  viz_dates <- seq.Date(as.Date("2024-01-01"), as.Date("2024-12-31"), by = "day")
  viz_pred <- L_eff_model$predict_fn(viz_dates, rep("weekday", length(viz_dates)))
  viz_pred_wkend <- L_eff_model$predict_fn(viz_dates, rep("weekend", length(viz_dates)))

  p_L_regression <- ggplot() +
    geom_ribbon(data = tibble(date = viz_dates, lo = viz_pred$L_mu * exp(-1.96 * viz_pred$L_sigma),
                              hi = viz_pred$L_mu * exp(1.96 * viz_pred$L_sigma)),
                aes(x = date, ymin = lo, ymax = hi), alpha = 0.15, fill = "steelblue") +
    geom_line(data = tibble(date = viz_dates, L = viz_pred$L_mu),
              aes(x = date, y = L, color = "Weekday prediction"), linewidth = 0.8) +
    geom_line(data = tibble(date = viz_dates, L = viz_pred_wkend$L_mu),
              aes(x = date, y = L, color = "Weekend prediction"), linewidth = 0.8, linetype = "dashed") +
    geom_point(data = L_eff_model$detail,
               aes(x = event_date, y = L_effective, shape = day_type_group), size = 3) +
    labs(title = "L_effective Regression (log-scale quadratic + day type)",
         subtitle = sprintf("Method: %s | n=%d I/E days | Shaded = 95%% prediction interval",
                            L_eff_model$method, L_eff_model$n_obs),
         x = "Date", y = "L_effective (hours)", color = "Prediction", shape = "I/E day type") +
    theme_bw()
  ggsave(file.path(output_dir, "plot_L_effective_regression.png"), p_L_regression, width = 12, height = 6)
}

# Day length comparison (civil twilight vs regression)
p_day_length <- days_full |>
  ggplot(aes(x = event_date)) +
  geom_line(aes(y = day_length_civil_twilight, color = "Civil twilight"), alpha = 0.5) +
  geom_ribbon(aes(ymin = L_mu * exp(-1.96 * L_prior_sigma),
                  ymax = L_mu * exp(1.96 * L_prior_sigma)),
              alpha = 0.15, fill = "steelblue") +
  geom_line(aes(y = L_mu, color = "I/E L_effective (regression)"), linewidth = 0.8) +
  labs(title = "Day Length with Uncertainty",
       subtitle = "Shaded band = 95% prior interval for L (propagated into Stan model)",
       x = "Date", y = "Day length (hours)", color = "Method") +
  scale_color_manual(values = c("Civil twilight" = "grey60", "I/E L_effective (regression)" = "steelblue")) +
  theme_bw()
ggsave(file.path(output_dir, "plot_day_length_comparison.png"), p_day_length, width = 10, height = 5)

cat("  Diagnostic plots saved.\n")
timer_stop("2_diagnostic_plots")


```

```{r input-figures-show, echo=TRUE, fig.height=6, fig.cap=c('Shore effort (gear counts) over the season by day type.','Private-boat effort (trailer counts) over the season by day type.','Interview CPUE over the season.','Effort by month.')}
for (nm in c('p_shore_effort','p_boat_effort','p_cpue','p_effort_month')) if (exists(nm)) print(get(nm))
```

## 3.5 Other-fishery opener spillover (diagnostic)

A hypothesis test, not a model change: do other coastal fishery openers pull effort or catch onto (or off of) crab days enough to deserve their own day category, the way crabbing holidays already do? Two openers are checked against the crab effort and CPUE series: Marine Area 2 finfish (salmon, halibut, bottomfish) from the MA2 opener calendar, and coastal razor-clam digs on the beaches nearest Grays Harbor (Twin Harbors, Copalis, Mocrocks) from the dig calendar. Because these openers cluster on weekends and in particular months, which already drive crab effort, the raw opener-vs-other comparison is shown alongside an **adjusted** association: the opener's coefficient in `lm(response ~ day_type + month + opener)`, i.e. the marginal effect after the weekend and monthly signal are removed. Read the adjusted column as the real test. This section changes no estimate; it only reports the associations so the analyst can decide whether to add an opener day category on a later run.

```{r fishery-spillover, results='hide', fig.show='hide'}
# ---- 3.5 OTHER-FISHERY OPENER SPILLOVER (diagnostic; changes no estimate) ----
spillover <- NULL; sp_raw_disp <- NULL; sp_adj_disp <- NULL; p_spill_eff <- NULL; p_spill_cpue <- NULL
if (isTRUE(params$run_fishery_spillover_diag)) {
  timer_start("2b_fishery_spillover")
  spillover <- tryCatch(
    diagnose_fishery_spillover(dwg, params),
    error = function(e) { cat(sprintf("  Fishery-opener spillover diagnostic skipped: %s\n", conditionMessage(e))); NULL })
  if (!is.null(spillover)) {
    write.csv(spillover$raw,      file.path(output_dir, "fishery_spillover_raw.csv"),      row.names = FALSE)
    write.csv(spillover$adjusted, file.path(output_dir, "fishery_spillover_adjusted.csv"), row.names = FALSE)
    for (nt in spillover$notes) cat("  NOTE:", nt, "\n")

    sp_raw_disp <- spillover$raw |>
      dplyr::transmute(Series, Opener,
        `n (opener)` = n_opener, `Mean (opener)` = round(mean_opener, 2),
        `n (other)`  = n_other,  `Mean (other)`  = round(mean_other, 2),
        `Raw diff`   = round(raw_diff, 2), `Raw %` = round(raw_pct, 0),
        `t-test p`   = round(t_p, 3))
    sp_adj_disp <- spillover$adjusted |>
      dplyr::transmute(Series, Opener,
        `Adj. estimate` = round(adj_estimate, 2), SE = round(se, 2),
        `95% CI` = ifelse(is.na(ci_lo), "-", sprintf("[%.1f, %.1f]", ci_lo, ci_hi)),
        p = round(adj_p, 3), Note = note)

    mk_long <- function(df) df |>
      tidyr::pivot_longer(dplyr::all_of(spillover$flag_cols), names_to = "opener", values_to = "is_open") |>
      dplyr::mutate(Opener = unname(spillover$flag_labels[opener]),
                    Status = ifelse(is_open, "Opener", "Other"))
    p_spill_eff <- ggplot(mk_long(spillover$effort_long), aes(Status, value, fill = Status)) +
      geom_boxplot(outlier.size = 0.5, alpha = 0.85) +
      facet_grid(series ~ Opener, scales = "free_y") +
      scale_fill_manual(values = c(Opener = "#c0392b", Other = "grey70")) +
      labs(title = "Crab effort: other-fishery opener days vs other days", x = NULL, y = "Daily effort count") +
      theme_bw() + theme(legend.position = "none", axis.text.x = element_text(angle = 20, hjust = 1))
    p_spill_cpue <- ggplot(mk_long(spillover$cpue_long), aes(Status, value, fill = Status)) +
      geom_boxplot(outlier.size = 0.5, alpha = 0.85) +
      facet_grid(series ~ Opener, scales = "free_y") +
      scale_fill_manual(values = c(Opener = "#c0392b", Other = "grey70")) +
      labs(title = "Interview CPUE: other-fishery opener days vs other days", x = NULL, y = "Dungeness per crabber-hour") +
      theme_bw() + theme(legend.position = "none", axis.text.x = element_text(angle = 20, hjust = 1))
  }
  timer_stop("2b_fishery_spillover")
}
```

```{r fishery-spillover-raw-show, echo=FALSE}
if (!is.null(sp_raw_disp)) {
  report_table(sp_raw_disp, paste("Raw comparison: mean crab effort (daily counts) and CPUE (Dungeness per",
    "crabber-hour) on other-fishery opener days vs other days. These are confounded with day-type and season;",
    "read the adjusted table below for the marginal effect."))
} else {
  cat(if (isTRUE(params$run_fishery_spillover_diag))
        "_Fishery-opener spillover diagnostic produced no output this run._"
      else "_Fishery-opener spillover diagnostic is off (`run_fishery_spillover_diag = FALSE`)._")
}
```

```{r fishery-spillover-adj-show, echo=FALSE}
if (!is.null(sp_adj_disp)) report_table(sp_adj_disp, paste("Adjusted association: the opener coefficient in",
  "lm(response ~ day-type + month + opener), the marginal effect after removing the weekend and monthly signal",
  "the model already captures. Positive = more effort/CPUE on opener days. A blank estimate with a Note means",
  "the opener could not be separated from day-type/month in this season."))
```

```{r fishery-spillover-notes, echo=FALSE, results='asis'}
if (!is.null(spillover)) {
  cat("**Identifiability notes (read before interpreting):**\n\n")
  for (nt in spillover$notes) cat("- ", nt, "\n\n", sep = "")
}
```

```{r fishery-spillover-plots, echo=FALSE, fig.height=7, fig.cap=c('Crab effort on other-fishery opener days vs other days, by opener and series.','Interview CPUE on other-fishery opener days vs other days, by opener and series.')}
if (!is.null(p_spill_eff))  print(p_spill_eff)
if (!is.null(p_spill_cpue)) print(p_spill_cpue)
```

### 3.6 Razor-dig shore-effort term (item 1)

Whether this run adds a `razor_dig` term to the SHORE effort model (analogous to the holiday term), controlled by `params$razor_dig_mode` (`"no"` | `"yes"` | `"auto"`). In `"auto"` it switches on only if the spillover diagnostic's day-type/month-adjusted `razor_nearby_dig` effect on shore effort is significant at `params$razor_dig_auto_p`. When active, the shore fits gain a `B3 * razor[d]` effort term (razor days from the consolidated opener calendar); to test it, compare the shore-effort `elpd_loo` (in `loo_summary_shore_*` / the LOO section) against a `razor_dig_mode = "no"` run. This changes the shore effort model, so it is off by default.

```{r razor-dig-decision, echo=TRUE}
# Nearby-beach dig dates from the consolidated opener calendar, and the on/off decision.
params$razor_dig_dates <- tryCatch({
  ev <- prep_fishery_events(params)
  ev$razor |> dplyr::filter(razor_nearby_dig) |> dplyr::pull(event_date)
}, error = function(e) as.Date(character(0)))

razor_mode   <- tolower(params$razor_dig_mode %||% "no")
razor_thresh <- params$razor_dig_auto_p %||% 0.05
razor_auto_p <- NA_real_
if (razor_mode == "auto" && !is.null(spillover)) {
  ap <- spillover$adjusted |>
    dplyr::filter(Series == "Shore gear (effort)", grepl("Razor", Opener))
  if (nrow(ap) > 0) razor_auto_p <- ap$adj_p[1]
}
params$razor_dig_active <- switch(razor_mode,
  yes  = TRUE,
  no   = FALSE,
  auto = is.finite(razor_auto_p) && razor_auto_p < razor_thresh,
  FALSE)
if (razor_mode == "auto" && is.null(spillover))
  cat("  NOTE: razor_dig_mode = 'auto' but the spillover diagnostic is off ",
      "(run_fishery_spillover_diag = FALSE); the razor term stays OFF.\n", sep = "")

cat(sprintf("Razor-dig shore-effort term: mode=%s, active=%s%s\n",
    razor_mode, params$razor_dig_active,
    if (razor_mode == "auto")
      sprintf(" (adjusted shore-effort p=%s, threshold=%.3f)",
              ifelse(is.finite(razor_auto_p), sprintf('%.3f', razor_auto_p), 'NA'), razor_thresh)
    else ""))
```

### 3.1 Ingress/egress and effective day length

The I/E surveys measure how long crabbers are actually present, which sets the effective day length used to expand snapshot effort counts. Civil twilight (9 to 17 hours) would overstate active fishing time by roughly a factor of two, so the I/E-derived length is used instead.

```{r ie-analysis, results='hide', fig.show='hide'}
# ---- 7.1b I/E ANALYSIS ----
cat("\n", strrep("=",60), "\n I/E ANALYSIS\n", strrep("=",60), "\n")

if(nrow(ie_data) > 0) {
  ie_validation <- ie_data |>
    filter(population == "shore") |>
    left_join(
      dwg$shore_effort |>
        group_by(event_date) |>
        summarise(mean_gear = mean(count_quantity), n_counts = n(), .groups = "drop"),
      by = "event_date"
    ) |>
    filter(!is.na(mean_gear), mean_gear > 0)

  if(nrow(ie_validation) > 0) {
    # Shared helper (03_R_functions/bss_day_length.R) so the [9, 17] cap lives in
    # exactly one place; this previously hardcoded a 16.0 cap.
    ie_validation$day_length_ct <- bss_day_length_civil(ie_validation$event_date, params)

    summ_shore_full <- prep_population_summary(dwg, "shore", params$est_date_start, params$est_date_end, params)
    cpg <- summ_shore_full$crabbers_per_gear

    ie_validation <- ie_validation |>
      mutate(
        gear_est_crabber_hrs = mean_gear * cpg * day_length_ct,
        f_temporal = ie_crabber_hours / gear_est_crabber_hrs,
        pct_error = (gear_est_crabber_hrs - ie_crabber_hours) / ie_crabber_hours * 100
      )

    cat("\n  I/E vs Gear-Count Comparison (paired days):\n")
    for(i in 1:nrow(ie_validation)) {
      r <- ie_validation[i,]
      cat(sprintf("  %-12s %-8s IE=%.0f Gear_est=%.0f f=%.2f\n",
                  as.character(r$event_date), r$day_type, r$ie_crabber_hours,
                  r$gear_est_crabber_hrs, r$f_temporal))
    }
    write.csv(ie_validation |>
      select(event_date, day_type, ie_crabber_hours, max_present, mean_present,
             L_effective, mean_gear, n_counts, gear_est_crabber_hrs, f_temporal, pct_error),
      file.path(output_dir, "ie_analysis.csv"), row.names = FALSE)
  }
} else {
  cat("  No I/E data available.\n")
}


```

```{r ie-analysis-show, echo=TRUE, fig.height=5}
for (nm in c('p_L_regression','p_day_length')) if (exists(nm)) print(get(nm))
```

### 3.2 Shore I/E representativeness (diagnostic)

The shore all-gear `sigma_IE` (~1.07 in the 2026-07-12 run, vs ~0.60 in the gear-resolved fit) means the I/E-derived effective day length and the effort counts disagree by roughly a factor of `e` at the tails. This check asks whether that is because the shore I/E days were sampled on unusually high-effort days (so the derived day length is unrepresentative) or is just sparse-data scatter across a handful of I/E days: it compares the shore daily gear counts on I/E observation days against all other sampled days (item 4 / GR-9).

```{r ie-representativeness, echo=FALSE}
ie_rep <- tryCatch(diagnose_ie_representativeness(dwg, ie_data, params), error = function(e) NULL)
if (!is.null(ie_rep)) {
  write.csv(ie_rep$table, file.path(output_dir, "ie_representativeness.csv"), row.names = FALSE)
  cat(sprintf("**Shore I/E days: %d.** %s\n\n", ie_rep$n_ie, ie_rep$verdict))
  report_table(ie_rep$table |> dplyr::mutate(dplyr::across(dplyr::where(is.numeric), ~round(., 1))),
    paste("Shore daily gear counts on I/E observation days vs other sampled days. A mean",
          "percentile rank near 0.5 with a non-significant rank test means the I/E days are",
          "representative, so a large shore sigma_IE reflects sparse-data scatter, not",
          "sampling on peak-effort days."))
} else {
  cat("_Shore I/E representativeness diagnostic unavailable (no shore I/E data)._")
}
```

## 4. Point Estimator (PE)

The design-based estimator: it averages sampled days within each stat-week by day-type stratum and expands to the unsampled days. The PE is fast and transparent, and it is the fallback whenever a Bayesian fit does not pass the convergence gate.

```{r pe-estimates, results='hide'}
# ---- 7.2 PE ESTIMATES ----
timer_start("3_PE_estimates")
cat("\n", strrep("=",60), "\n PE ESTIMATES\n", strrep("=",60), "\n")

pe_all <- list()
for(pop in c("shore","private_boat")) {
  for(ss in subseasons) {
    label <- paste0(pop, "_", ss$name)
    days_ss <- prep_days_crab(ss$start, ss$end, params, L_eff_model = L_eff_model)
    summ_ss <- prep_population_summary(dwg, pop, ss$start, ss$end, params)
    pe_all[[label]] <- run_pe_pooled(summ_ss, days_ss, params, label)
  }
}
timer_stop("3_PE_estimates")


```

### 4.1 Incomplete-trip sensitivity

Incomplete trips (gear not yet retrieved at the interview) read systematically low for pots and traps. This section quantifies the harvest difference between keeping and dropping them; the production default drops them (`filter_incomplete_trips = TRUE`).

```{r pe-sensitivity, results='hide'}
# ---- 7.2b POOL-2: INCOMPLETE-TRIP SENSITIVITY ----
# Quantify the harvest impact of the incomplete-trip filter: re-run the fast,
# design-based PE estimator with the filter OFF vs ON for each shore/boat
# sub-season and report the % change in Dungeness catch. Incomplete trips read
# ~-20% low for soak-time gear, so the filtered (complete-only) estimate is
# expected to be higher. This section is DIAGNOSTIC ONLY: it re-runs PE (no MCMC),
# changes no reported total, and is tryCatch-wrapped so it can never break the
# run. Output: sensitivity_incomplete_trips.csv.
cat("\n", strrep("=",60), "\n INCOMPLETE-TRIP SENSITIVITY (POOL-2)\n", strrep("=",60), "\n")
sens_df <- NULL   # persists to the display chunk below even if the tryCatch errors
tryCatch({
  sens_rows <- list()
  for(pop in c("shore","private_boat")) {
    for(ss in subseasons) {
      s_label <- paste0(pop, "_", ss$name)
      days_ss <- prep_days_crab(ss$start, ss$end, params, L_eff_model = L_eff_model)
      summ_ss <- prep_population_summary(dwg, pop, ss$start, ss$end, params)
      p_off   <- modifyList(params, list(filter_incomplete_trips = FALSE))
      p_on    <- modifyList(params, list(filter_incomplete_trips = TRUE))
      d_off   <- run_pe_pooled(summ_ss, days_ss, p_off, paste0(s_label, " [all trips]"))$Dungeness_Kept %||% 0
      d_on    <- run_pe_pooled(summ_ss, days_ss, p_on,  paste0(s_label, " [complete only]"))$Dungeness_Kept %||% 0
      sens_rows[[s_label]] <- tibble(
        component                  = s_label,
        component_disp             = paste0(gsub("_", " ", pop), " ", ss$display_name),
        PE_Dungeness_all_trips     = round(d_off),
        PE_Dungeness_complete_only = round(d_on),
        diff_pct                   = if(d_off > 0) round((d_on - d_off) / d_off * 100, 1) else NA_real_
      )
    }
  }
  sens_df <- bind_rows(sens_rows)
  sens_df$filter_applied_in_run <- isTRUE(params$filter_incomplete_trips)
  write.csv(sens_df, file.path(output_dir, "sensitivity_incomplete_trips.csv"), row.names = FALSE)
}, error = function(e) cat(sprintf("  Incomplete-trip sensitivity skipped: %s\n", conditionMessage(e))))


```

```{r pe-sensitivity-show, echo=FALSE}
# Display the incomplete-trip sensitivity so the section shows a result on the page
# (previously it only wrote a CSV). Positive "Change (%)" is the upward CPUE/harvest
# correction from dropping the low-biased incomplete trips (the production default).
if (!is.null(sens_df) && nrow(sens_df) > 0) {
  report_table(
    dplyr::transmute(sens_df,
      Component                 = component_disp,
      `PE catch, all trips`     = PE_Dungeness_all_trips,
      `PE catch, complete only` = PE_Dungeness_complete_only,
      `Change (%)`              = diff_pct,
      `Filter applied in run`   = filter_applied_in_run),
    caption = paste("Incomplete-trip filter sensitivity: design-based (PE) Dungeness catch by",
                    "component and sub-season, all trips vs complete-only. The production run drops",
                    "incomplete trips (Filter applied in run = TRUE); a positive Change (%) is the",
                    "upward correction from removing the low-biased incomplete trips. Diagnostic only."))
} else {
  cat("_Incomplete-trip sensitivity produced no output this run._")
}
```

### 4.2 Commercial and charter census

The commercial and charter harvest is a day-type-stratified census expansion of the daily vessel tally, not a modeled quantity. As of the T1.4 change it uses a **separate mean catch per vessel for commercial vs charter vessels**, each applied to its own tally column, rather than one mean across the pooled tally; the two classes have different catch profiles, so a single mean biased the total whenever the commercial-to-charter mix in the tally differed from the mix in the interviews.

```{r pe-comm, results='hide'}
# ---- 7.4 COMMERCIAL/CHARTER ----
timer_start("4_commercial_charter")
pe_all$comm_charter <- estimate_comm_charter(dwg, params)
timer_stop("4_commercial_charter")


```

```{r pe-comm-show, echo=FALSE}
# T1.4: show the commercial-vs-charter per-vessel split behind the census. Each class
# gets its own mean Dungeness per vessel applied to its own tally column (Guide folded
# into Charter); a class with no interviews falls back to the pooled mean.
.vt <- tryCatch(pe_all$comm_charter$vessel_type_detail, error = function(e) NULL)
if (!is.null(.vt) && nrow(.vt) > 0) {
  report_table(
    dplyr::transmute(.vt,
      `Vessel type`            = vessel_type,
      Interviews               = n_interviews,
      `Mean Dungeness/vessel`  = round(mean_dung_per_vessel, 1),
      `Tally vessels (window)` = round(tally_vessels),
      `Sampled-day catch`      = round(sampled_catch)),
    caption = paste("Commercial vs charter split behind the census (T1.4). Each class's",
                    "mean catch per vessel is applied to its own tally column instead of",
                    "one pooled mean. 'Sampled-day catch' is the class census before the",
                    "day-type expansion that produces the reported commercial/charter",
                    "total, so it shows the split, not the final stratified figure."))
} else {
  cat("_Commercial/charter vessel-type split unavailable this run._")
}
```

### 4.3 PE port totals

The Point Estimator port total, summed across the three population components. Compare it against the Bayesian total below.

```{r pe-port, results='hide'}
# ---- 7.5 PE PORT TOTALS ----
cat("\n", strrep("=",60), "\n PE PORT TOTALS\n", strrep("=",60), "\n")

# component_keys / component_names now follow the sub-season set (build_subseasons):
# one row per population x sub-season, then comm_charter, then a PORT TOTAL computed from
# the semantic sums (pe_port_total), not by a fixed row index. component_keys is still
# consumed by the PE-consistency check chunk below (comm_charter has no check, skipped there).
pe_pops    <- c("shore", "private_boat")
pop_disp   <- c(shore = "Shore", private_boat = "Private Boat")
ss_names   <- vapply(subseasons, function(ss) ss$name,         character(1))
ss_disp    <- vapply(subseasons, function(ss) ss$display_name, character(1))
component_keys  <- c(unlist(lapply(pe_pops, function(p) paste0(p, "_", ss_names))), "comm_charter")
component_names <- c(unlist(lapply(pe_pops, function(p) sprintf("%s (%s)", pop_disp[[p]], ss_disp))),
                     "Commercial/Charter")

pe_summary <- tibble(
  Component  = c(component_names, "PORT TOTAL"),
  Effort_hrs = c(vapply(component_keys, function(k) pe_field(k, "effort_total"),  numeric(1)), pe_port_total("effort_total")),
  Dungeness  = c(vapply(component_keys, function(k) pe_field(k, "Dungeness_Kept"), numeric(1)), pe_port_total("Dungeness_Kept"))
)
if(params$estimate_red_rock) {
  pe_summary$Red_Rock <- c(vapply(component_keys, function(k) pe_field(k, "Red_Rock_Kept"), numeric(1)),
                           pe_port_total("Red_Rock_Kept"))
}

print(pe_summary |> mutate(across(where(is.numeric), ~round(.))))
write.csv(pe_summary, file.path(output_dir, "pe_port_summary.csv"), row.names=FALSE)


```

```{r pe-port-show, echo=TRUE}
report_table(pe_summary, 'Point Estimator (PE) port totals by component.')
```

### 4.4 PE internal consistency (ratio-of-sums)

Each component's PE stratum CPUE is a ratio-of-sums (`sum(catch)/sum(hrs)` within each week-by-day-type stratum), not a mean of per-day ratios (backlog P0). The mean-of-ratios was unstable in thin strata, a day with very little sampled effort produced an extreme daily ratio that was then multiplied by the full stratum effort, and it made the shore PE's implied CPUE about 2.9x its own interview ratio-of-sums. Ratio-of-sums removes that and matches the gear-resolved track. The check below is the guardrail: the PE's implied CPUE (its catch divided by its effort) must agree with the ratio-of-sums over the same interviews. A ratio far from 1 means catch and effort are on different scales, and the run stops if any component exceeds 2x, so a clean table here is itself the confirmation that the PE is internally consistent.

```{r pe-consistency-show, echo=FALSE}
# P0: render the PE internal-consistency check returned by run_pe_pooled per component
# (component_keys is defined in the pe-port chunk above). comm_charter is a census, not
# run_pe_pooled, so it has no check and is skipped.
pe_chk <- tryCatch(
  dplyr::bind_rows(lapply(component_keys, function(k) {
    d <- pe_all[[k]]$pe_cpue_check
    if (is.null(d)) NULL else dplyr::mutate(d, Component = k)
  })),
  error = function(e) NULL)
if (!is.null(pe_chk) && nrow(pe_chk) > 0) {
  report_table(
    dplyr::transmute(pe_chk,
      Component         = short_fit(Component),
      Unit              = effort_unit,
      `PE implied CPUE` = round(pe_implied_cpue, 3),
      `Interview RoS`   = round(interview_ros, 3),
      Ratio             = round(ratio, 2)),
    caption = paste("PE internal-consistency check (P0). 'PE implied CPUE' is the",
                    "component's PE catch / PE effort; 'Interview RoS' is the ratio-of-sums",
                    "over the same interviews. Healthy: Ratio near 1 (the run stops if any",
                    "component is below 0.5 or above 2.0)."))
} else {
  cat("_PE consistency check unavailable (no fitted components produced it)._")
}
```

## 5. Bayesian state-space model fits

Each population by sub-season is fit independently with a Bayesian state-space (BSS) model that treats daily effort and catch rate as smoothly evolving latent series, fills the gaps between sampled days, and returns full credible intervals. **What a healthy fit looks like:** R-hat below 1.01, effective sample size above 400, and few or no divergent transitions on every monitored quantity. Fits are checked against these criteria in the next section.

```{r bss-fits, results='hide', fig.show='hide'}
# ---- 7.6 BSS MODEL RUNS ----
timer_start("5_BSS_model_runs")
cat("\n", strrep("=",60), "\n BSS MODEL RUNS (Adaptive AR)\n", strrep("=",60), "\n")

bss_catch_groups <- catch_groups
bss_all <- list()
ar_resolutions <- list()  # track AR resolution per fit

for(cg in bss_catch_groups) {
  for(pop in c("shore","private_boat")) {
    for(ss in subseasons) {
      label <- paste0(pop, "_", ss$name, "_", cg)
      cat(sprintf("\n--- BSS: %s ---\n", label))

      days_ss <- prep_days_crab(ss$start, ss$end, params, L_eff_model = L_eff_model)
      summ_ss <- prep_population_summary(dwg, pop, ss$start, ss$end, params)

      # NOTE (POOL-2): n_int is the UNFILTERED interview count. The fitted CPUE
      # count is lower once prep_bss_crab drops h<=0 rows and (with
      # filter_incomplete_trips = TRUE) incomplete trips, so this sufficiency
      # guard reads looser than the count the fit actually uses. Behavior is
      # unchanged from prior versions (the guard was always pre-filter); on the
      # 2024-25 data every fitted component keeps well over bss_min_interviews
      # complete interviews. If a future component is marginal, gate on the
      # complete-only count instead.
      n_int <- nrow(summ_ss$interview)
      n_eff <- nrow(summ_ss$effort_index)

      if(n_eff < 5 || n_int < params$bss_min_interviews) {
        cat(sprintf("  Insufficient data (%d effort, %d interviews), PE fallback.\n", n_eff, n_int))
        bss_all[[label]] <- list(C_sum=NA, E_sum=NA, pe_fallback=TRUE,
                                 population=pop, subseason=ss$name, catch_group=cg)
        next
      }

      bss_data <- prep_bss_crab_pooled(days_ss, summ_ss, cg, params, pop, ie_data = ie_data)
      ar_resolutions[[label]] <- attr(bss_data, "ar_resolution")

      is_shore   <- (pop == "shore")
      is_boat    <- (pop == "private_boat")
      is_allgear <- (ss$gear_regime == "all_gear")
      is_ringnet <- (ss$gear_regime == "pot_closure")

      # v6.1/v6.2: per-fit sampler tuning. Each branch sets iter, warmup,
      # max_treedepth, and adapt_delta; any fit without a specific override
      # uses the *_default values. Ported from the gear-resolved track (v6.1)
      # and extended with a boat all-gear branch (v6.2). The boat all-gear fit
      # produced 5998 divergences with treedepth_pct = 0 in the 2026-04-08 run,
      # so step size (adapt_delta), not trajectory length, was the binding
      # constraint: adapt_delta is raised to 0.99 and max_treedepth to 13 (the
      # smaller steps from a higher adapt_delta lengthen trajectories, so the
      # deeper tree is a buffer against new treedepth saturation).
      if(is_shore && is_allgear) {
        fit_iter    <- params$bss_iter_shore_allgear
        fit_warmup  <- params$bss_warmup_shore_allgear
        fit_treedep <- params$bss_treedepth_shore_allgear %||% params$bss_max_treedepth_default
        fit_delta   <- params$bss_delta_shore_allgear     %||% params$bss_adapt_delta
      } else if(is_boat && is_allgear) {
        fit_iter    <- params$bss_iter_boat_allgear      %||% params$bss_iter_default
        fit_warmup  <- params$bss_warmup_boat_allgear    %||% params$bss_warmup_default
        fit_treedep <- params$bss_treedepth_boat_allgear %||% params$bss_max_treedepth_default
        fit_delta   <- params$bss_delta_boat_allgear     %||% params$bss_adapt_delta
      } else if(is_ringnet) {
        fit_iter    <- params$bss_iter_default
        fit_warmup  <- params$bss_warmup_default
        fit_treedep <- params$bss_treedepth_ringnet %||% params$bss_max_treedepth_default
        fit_delta   <- params$bss_delta_ringnet     %||% params$bss_adapt_delta
      } else {
        fit_iter    <- params$bss_iter_default
        fit_warmup  <- params$bss_warmup_default
        fit_treedep <- params$bss_max_treedepth_default
        fit_delta   <- params$bss_adapt_delta
      }

      cat(sprintf("  Stan: iter=%d, warmup=%d, max_treedepth=%d, adapt_delta=%.2f, D=%d days, P_n=%d (%s), %d interviews\n",
                  fit_iter, fit_warmup, fit_treedep, fit_delta, bss_data$D, bss_data$P_n,
                  attr(bss_data, "ar_resolution"), bss_data$IntC))

      timer_start(paste0("BSS_", label))
      fit <- rstan::stan(
        file = here("02_stan_models", params$bss_model_file),
        data = bss_data,
        chains = params$bss_chains, iter = fit_iter, warmup = fit_warmup,
        control = list(adapt_delta=fit_delta, max_treedepth=fit_treedep),
        seed = params$bss_seed,
        cores = params$bss_cores
      )
      timer_stop(paste0("BSS_", label))

      # --- Extract BOTH expected and predictive catch ---
      bss_pars <- c("C_sum","C_expected_sum","E_sum","B1","B2","B1_C_out","B2_C_out")
      if(bss_data$IE_n > 0) bss_pars <- c(bss_pars, "sigma_IE_out")
      if(isTRUE(params$razor_dig_active) && pop == "shore") bss_pars <- c(bss_pars, "B3_out")   # item 1
      if(isTRUE(params$estimate_cpue_density)) bss_pars <- c(bss_pars, "gamma_C_out")           # item 6b
      bss_summ <- summary(fit, pars=bss_pars)$summary
      cat("  BSS Summary:\n"); print(bss_summ)
      write.csv(as.data.frame(bss_summ), file.path(output_dir, paste0("bss_summary_",label,".csv")))

      C_draws <- rstan::extract(fit, "C_sum")$C_sum           # predictive (with Poisson noise)
      C_exp_draws <- rstan::extract(fit, "C_expected_sum")$C_expected_sum  # expected (no noise)
      E_draws <- rstan::extract(fit, "E_sum")$E_sum

      bss_all[[label]] <- list(
        C_sum = quantile(C_draws, c(0.025,0.5,0.975)),
        C_expected_sum = quantile(C_exp_draws, c(0.025,0.5,0.975)),
        E_sum = quantile(E_draws, c(0.025,0.5,0.975)),
        C_draws = C_draws, C_exp_draws = C_exp_draws, E_draws = E_draws,
        fit = fit, pe_fallback = FALSE, bss_data = bss_data, days_ss = days_ss,
        population = pop, subseason = ss$name, catch_group = cg
      )

      # --- Use C_expected (no Poisson noise) for primary estimates ---
      E_bss <- rstan::extract(fit, "E")$E
      C_exp_bss <- rstan::extract(fit, "C_expected")$C_expected

      E_daily <- tibble(
        event_date = days_ss$event_date,
        median = apply(E_bss[,1,,1], 2, median),
        lo95 = apply(E_bss[,1,,1], 2, quantile, 0.025),
        hi95 = apply(E_bss[,1,,1], 2, quantile, 0.975)
      )

      C_daily <- tibble(
        event_date = days_ss$event_date,
        median = apply(C_exp_bss[,1,,1], 2, median),
        lo95 = apply(C_exp_bss[,1,,1], 2, quantile, 0.025),
        hi95 = apply(C_exp_bss[,1,,1], 2, quantile, 0.975)
      )

      # BSS effort plot
      p_bss_eff <- E_daily |>
        ggplot(aes(x = event_date)) +
        geom_ribbon(aes(ymin = lo95, ymax = hi95), alpha = 0.2, fill = "steelblue") +
        geom_line(aes(y = median), color = "steelblue") +
        labs(title = sprintf("BSS Daily Effort: %s", label),
             subtitle = "Median with 95% credible interval (daily AR(1))",
             x = "Date", y = "Crabber-hours") +
        theme_bw()
      ggsave(file.path(output_dir, paste0("plot_bss_effort_", label, ".png")),
             p_bss_eff, width = 10, height = 6)

      # BSS catch plot (using expected catch)
      p_bss_catch <- C_daily |>
        ggplot(aes(x = event_date)) +
        geom_ribbon(aes(ymin = lo95, ymax = hi95), alpha = 0.2, fill = "darkgreen") +
        geom_line(aes(y = median), color = "darkgreen") +
        labs(title = sprintf("BSS Daily Expected Catch: %s", label),
             subtitle = "E[C|data] with 95% CI (no Poisson sampling noise)",
             x = "Date", y = "Expected Dungeness caught") +
        theme_bw()
      ggsave(file.path(output_dir, paste0("plot_bss_catch_", label, ".png")),
             p_bss_catch, width = 10, height = 6)

      # Posterior densities
      p_bss_post <- tibble(
        `Expected Catch` = C_exp_draws,
        `Predictive Catch` = C_draws,
        `Effort` = E_draws
      ) |>
        pivot_longer(everything(), names_to = "param", values_to = "value") |>
        ggplot(aes(x = value, fill = param)) +
        geom_density(alpha = 0.4) +
        facet_wrap(~param, scales = "free") +
        labs(title = sprintf("BSS Posteriors: %s", label), x = "Value", y = "Density") +
        theme_bw() + theme(legend.position = "none")
      ggsave(file.path(output_dir, paste0("plot_bss_posteriors_", label, ".png")),
             p_bss_post, width = 14, height = 5)

      write.csv(E_daily, file.path(output_dir, paste0("bss_daily_effort_", label, ".csv")), row.names=FALSE)
      write.csv(C_daily, file.path(output_dir, paste0("bss_daily_catch_", label, ".csv")), row.names=FALSE)

      # Save L posterior if estimated
      if(bss_data$estimate_L == 1) {
        L_post <- rstan::extract(fit, "L_out")$L_out  # [draws, D]
        L_daily <- tibble(
          event_date = days_ss$event_date,
          L_prior_mu = days_ss$L_mu,
          L_posterior_median = apply(L_post, 2, median),
          L_posterior_lo95 = apply(L_post, 2, quantile, 0.025),
          L_posterior_hi95 = apply(L_post, 2, quantile, 0.975)
        )
        write.csv(L_daily, file.path(output_dir, paste0("bss_L_effective_", label, ".csv")), row.names=FALSE)
      }
    }
  }
}
timer_stop("5_BSS_model_runs")


```

## 6. Convergence gate and model selection

The scale-aware gate decides, per fit, whether to report the Bayesian posterior or fall back to the PE. A fit passes only if R-hat is below 1.01, effective sample size is above 400, the divergent fraction is under the 15 percent backstop, and the divergent draws do not move the reported totals by more than 0.10 posterior standard deviations. **If a component reads PE here, its Bayesian fit did not clear the gate;** that is expected for data-thin components (historically the private boat) and is not an error.

```{r gate, results='hide'}
# ---- 7.6b SCALE-AWARE CONVERGENCE GATE (B1.8) ----
# Compute the pass/fail gate ONCE per fit here, immediately after fitting, and
# store the decision (use_bss) plus a full diagnostic row (gate_info) on each
# bss_all entry. Sections 7.7, 7.8, and 7.8b consume use_bss through bss_use_pe_for()
# below, and 7.11 simply assembles the stored gate_info rows. This makes the gate
# the single authoritative source of method selection: previously 7.7/7.8/7.8b
# keyed off pe_fallback (data sufficiency) alone, so a fit with enough data but a
# failed gate would still have been summed into the BSS totals. Wiring use_bss
# into the totals closes that gap. On the 2024-25 data all three fitted
# components pass the scale-aware gate, so the port total is unchanged; but a
# future weakly-identified fit that fails the gate now correctly falls back to PE
# everywhere, not only in the report labels.
cat("\n", strrep("=",60), "\n CONVERGENCE GATE (scale-aware, B1.8)\n", strrep("=",60), "\n")

# The scale-aware gate lives in the shared module
# 03_R_functions/bss_convergence_gate.R (bss_compute_gate / bss_use_pe_for), so the
# pooled and gear-resolved tracks share ONE gate and cannot drift. bss_compute_gate()
# takes the two gate thresholds as arguments (pooled passes its own params values)
# and the per-fit AR-resolution label for the report column; it gates on
# C_expected_sum and E_sum. The single
# pooled-only, REPORT-ONLY column B1_C_rhat (weekend-CPUE-effect R-hat, never part
# of the pass/fail decision) is re-attached at the call site so
# convergence_report.csv keeps its historical schema.
for(label in names(bss_all)) {
  b  <- bss_all[[label]]
  gi <- bss_compute_gate(
    b, label,
    catch_par        = "C_expected_sum",
    effort_par       = "E_sum",
    max_impact_sd    = params$max_divergence_total_impact_sd %||% 0.10,
    # Fallback matches the params value set above (0.05 as of P7, 2026-07-12) and the
    # shared function default, so the gate behaves identically if the param is ever
    # absent (pooled passes params$max_divergence_fraction = 0.05).
    max_div_fraction = params$max_divergence_fraction %||% 0.05,
    ar_resolution    = ar_resolutions[[label]] %||% NA_character_
  )
  # Re-attach the pooled-only B1_C_rhat report column (report-only, NOT gating),
  # right after E_sum_neff to preserve the historical column order. tryCatch
  # yields NA for PE-only entries (no fit to summarize).
  b1c_rhat <- tryCatch(
    round(summary(b$fit, pars = "B1_C_out")$summary["B1_C_out", "Rhat"], 4),
    error = function(e) NA_real_)
  gi <- tibble::add_column(gi, B1_C_rhat = b1c_rhat, .after = "E_sum_neff")
  bss_all[[label]]$gate_info <- gi
  bss_all[[label]]$use_bss   <- isTRUE(gi$pass_convergence)
}

# PE-vs-BSS selection uses the shared bss_use_pe_for(): report PE if the fit never
# ran OR failed the gate. Sections 7.7 / 7.8 / 7.8b call bss_use_pe_for() below.


```

## 7. Headline result: port total

The season harvest estimate: the sum of all components, reported as expected catch with a 95 percent credible interval. This is the number to cite, read together with the convergence and PE-vs-BSS sections. Components that passed the gate contribute their Bayesian posterior; components that did not contribute their PE point.

```{r port-total, results='hide'}
# ---- 7.7 COMBINED BSS PORT TOTAL ----
timer_start("6_outputs")
cat("\n", strrep("=",60), "\n COMBINED RESULTS\n", strrep("=",60), "\n")

n_draws_max <- max(sapply(bss_all, function(b) {
  if(bss_use_pe_for(b)) return(0)   # B1.8: a gate-failed fit contributes no BSS draws
  length(b$C_exp_draws)
}))
if(n_draws_max == 0) n_draws_max <- params$bss_chains * (params$bss_iter_default - params$bss_warmup_default)

for(cg in bss_catch_groups) {
  bss_C_total <- rep(0, n_draws_max)    # expected catch
  bss_C_pred_total <- rep(0, n_draws_max) # predictive catch
  bss_E_total <- rep(0, n_draws_max)

  for(pop in c("shore","private_boat")) {
    for(ss in subseasons) {
      label <- paste0(pop, "_", ss$name, "_", cg)
      pe_label <- paste0(pop, "_", ss$name)
      b <- bss_all[[label]]

      if(bss_use_pe_for(b)) {   # B1.8: PE if insufficient data OR failed scale-aware gate
        bss_C_total <- bss_C_total + (pe_all[[pe_label]][[cg]] %||% 0)
        bss_C_pred_total <- bss_C_pred_total + (pe_all[[pe_label]][[cg]] %||% 0)
        bss_E_total <- bss_E_total + pe_all[[pe_label]]$effort_total
      } else {
        n <- length(b$C_exp_draws)
        idx <- rep_len(1:n, n_draws_max)
        bss_C_total <- bss_C_total + b$C_exp_draws[idx]
        bss_C_pred_total <- bss_C_pred_total + b$C_draws[idx]
        bss_E_total <- bss_E_total + b$E_draws[idx]
      }
    }
  }
  bss_C_total <- bss_C_total + (pe_all$comm_charter[[cg]] %||% 0)
  bss_C_pred_total <- bss_C_pred_total + (pe_all$comm_charter[[cg]] %||% 0)
  bss_E_total <- bss_E_total + pe_all$comm_charter$effort_total

  final_table <- tibble(
    Catch_Group = cg,
    Estimate = c("Effort", "Expected_Catch", "Predictive_Catch"),
    PE = c(round(pe_port_total("effort_total")),
           round(if(cg=="Dungeness_Kept") pe_port_total("Dungeness_Kept") else pe_port_total("Red_Rock_Kept")),
           NA),
    BSS_median = c(round(median(bss_E_total)), round(median(bss_C_total)), round(median(bss_C_pred_total))),
    BSS_lo95 = c(round(quantile(bss_E_total,0.025)), round(quantile(bss_C_total,0.025)), round(quantile(bss_C_pred_total,0.025))),
    BSS_hi95 = c(round(quantile(bss_E_total,0.975)), round(quantile(bss_C_total,0.975)), round(quantile(bss_C_pred_total,0.975)))
  )

  cat(sprintf("\n--- Port Total: %s ---\n", cg))
  print(final_table)
  write.csv(final_table, file.path(output_dir, paste0("port_total_",cg,".csv")), row.names=FALSE)
}


```

```{r port-total-show, echo=TRUE}
report_table(final_table, 'Port total harvest estimate (expected catch with 95% credible interval), by component and combined.')
```

## 8. Monthly estimates

The season total broken out by month, with per-month credible intervals, showing the within-season harvest trajectory.

```{r monthly, results='hide', fig.show='hide'}
# ---- 7.8 MONTHLY ESTIMATES ----
cat("\n", strrep("=",60), "\n MONTHLY ESTIMATES\n", strrep("=",60), "\n")

days_full <- prep_days_crab(params$est_date_start, params$est_date_end,
                            params, L_eff_model = L_eff_model)

month_lookup <- days_full |>
  mutate(month_label = format(event_date, "%Y-%m")) |>
  select(event_date, month_label)

all_months <- sort(unique(month_lookup$month_label))
monthly_C_draws <- matrix(0, nrow = n_draws_max, ncol = length(all_months),
                          dimnames = list(NULL, all_months))
monthly_E_draws <- matrix(0, nrow = n_draws_max, ncol = length(all_months),
                          dimnames = list(NULL, all_months))

for(cg in bss_catch_groups) {
  for(pop in c("shore","private_boat")) {
    for(ss in subseasons) {
      label <- paste0(pop, "_", ss$name, "_", cg)
      pe_label <- paste0(pop, "_", ss$name)
      b <- bss_all[[label]]

      days_ss <- prep_days_crab(ss$start, ss$end, params, L_eff_model = L_eff_model)
      days_ss$month_label <- format(days_ss$event_date, "%Y-%m")

      if(bss_use_pe_for(b)) {   # B1.8: PE if insufficient data OR failed scale-aware gate
        summ_ss <- prep_population_summary(dwg, pop, ss$start, ss$end, params)
        # Fix 2 (v7.0), POOL-3 (v7.6): population-aware daily effort so the monthly
        # PE SHARE uses the same effort definition run_pe_pooled() uses. run_pe_pooled() builds the
        # private-boat effort series on the gear-DEPLOYMENT scale,
        # mean_count * gear_per_group * tau_boat (day-length-free), and shore as
        # crabber-hours, mean_count * crabbers_per_gear * day_length. The share is
        # normalized, so component and port TOTALS are unaffected; only the across-
        # month distribution changes. gear_per_group is recomputed here from the same
        # sub-season interview filter run_pe_pooled() uses so the two stay consistent.
        # T4.4: shared monthly effort-share helper (03_R_functions/pe_monthly_effort_share.R).
        monthly_pe <- pe_monthly_effort_share(pop, summ_ss, days_ss, params)

        pe_catch_total <- pe_all[[pe_label]][[cg]] %||% 0
        pe_effort_total <- pe_all[[pe_label]]$effort_total

        for(m in monthly_pe$month_label) {
          if(m %in% all_months) {
            sh <- monthly_pe$share[monthly_pe$month_label == m]
            monthly_C_draws[, m] <- monthly_C_draws[, m] + pe_catch_total * sh
            monthly_E_draws[, m] <- monthly_E_draws[, m] + pe_effort_total * sh
          }
        }
      } else {
        # Use C_expected (no Poisson noise) for monthly totals
        C_exp_bss <- rstan::extract(b$fit, "C_expected")$C_expected
        E_bss <- rstan::extract(b$fit, "E")$E
        n_fit_draws <- dim(C_exp_bss)[1]

        C_daily_mat <- C_exp_bss[,1,,1]
        E_daily_mat <- E_bss[,1,,1]

        for(m in unique(days_ss$month_label)) {
          day_idx <- which(days_ss$month_label == m)
          if(length(day_idx) == 0) next
          month_C <- if(length(day_idx)==1) C_daily_mat[, day_idx] else rowSums(C_daily_mat[, day_idx])
          month_E <- if(length(day_idx)==1) E_daily_mat[, day_idx] else rowSums(E_daily_mat[, day_idx])
          idx <- rep_len(1:n_fit_draws, n_draws_max)
          if(m %in% all_months) {
            monthly_C_draws[, m] <- monthly_C_draws[, m] + month_C[idx]
            monthly_E_draws[, m] <- monthly_E_draws[, m] + month_E[idx]
          }
        }
      }
    }
  }
}

# Add commercial/charter
if(!is.null(pe_all$comm_charter$daily_est)) {
  comm_daily <- pe_all$comm_charter$daily_est |>
    mutate(month_label = format(date, "%Y-%m"))
  comm_total <- pe_all$comm_charter$Dungeness_Kept
  total_sampled_dung <- sum(comm_daily$est_dung)
  if(total_sampled_dung > 0) {
    for(m in unique(comm_daily$month_label)) {
      if(m %in% all_months) {
        share <- sum(comm_daily$est_dung[comm_daily$month_label == m]) / total_sampled_dung
        monthly_C_draws[, m] <- monthly_C_draws[, m] + comm_total * share
      }
    }
  }
}

monthly_summary <- tibble(
  Month = all_months,
  BSS_Catch_median = apply(monthly_C_draws, 2, median),
  BSS_Catch_lo95   = apply(monthly_C_draws, 2, quantile, 0.025),
  BSS_Catch_hi95   = apply(monthly_C_draws, 2, quantile, 0.975),
  BSS_Effort_median = apply(monthly_E_draws, 2, median),
  BSS_Effort_lo95   = apply(monthly_E_draws, 2, quantile, 0.025),
  BSS_Effort_hi95   = apply(monthly_E_draws, 2, quantile, 0.975)
) |>
  mutate(across(where(is.numeric), round))

print(monthly_summary |> select(Month, BSS_Catch_median, BSS_Catch_lo95, BSS_Catch_hi95))
write.csv(monthly_summary, file.path(output_dir, "monthly_estimates.csv"), row.names = FALSE)

p_monthly <- monthly_summary |>
  mutate(Month = factor(Month, levels = all_months)) |>
  ggplot(aes(x = Month)) +
  geom_col(aes(y = BSS_Catch_median), fill = "steelblue", alpha = 0.7) +
  geom_errorbar(aes(ymin = BSS_Catch_lo95, ymax = BSS_Catch_hi95), width = 0.3) +
  labs(title = "Monthly Dungeness Catch (Expected, no Poisson noise)",
       x = "Month", y = "Dungeness crab caught") +
  theme_bw() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
ggsave(file.path(output_dir, "plot_monthly_catch.png"), p_monthly, width = 12, height = 6)


# ===========================================================================
```

```{r monthly-show, echo=TRUE, fig.height=6, fig.cap='Monthly Dungeness catch with 95% intervals.'}
report_table(monthly_summary, 'Monthly harvest estimates with 95% intervals.'); if (exists('p_monthly')) print(p_monthly)
```

### 8.1 Monthly catch by crabbing mode

The monthly trajectory split by mode (shore, private boat, commercial/charter). The uncertainty source matches the method selected for each component, so the boat interval here is a lower bound when the boat is on PE.

```{r monthly-mode, results='hide', fig.show='hide'}
# 7.8b MONTHLY ESTIMATES BY CRABBING MODE (with per-mode uncertainty)
# ===========================================================================
# Companion to plot_monthly_catch.png. Breaks the monthly Dungeness catch out by
# crabbing mode, each month with a 95% interval. The uncertainty source matches
# the method actually selected for each component (convergence_report / pe_vs_bss):
#   Shore  (BSS, both sub-seasons): full posterior. Monthly catch draws are summed
#          from the daily C_expected draws, then 2.5/50/97.5 quantiles.
#   Private Boat: method is per sub-season (v7.0 scale-aware gate, B1.8). On the
#          2024-25 data the all-gear sub-season passes the gate and uses the BSS
#          posterior (full daily C_expected draws, CV ~27%), while the pot-closure
#          sub-season has too few interviews (< bss_min_interviews) and uses PE.
#          The PE branch distributes the selected PE point across months by monthly
#          effort share, then applies a design-based relative SE (stratified effort
#          SE / effort total) as ONE per-component lognormal scale shared across
#          months, so the monthly catches move together and the boat total carries
#          that relative SE. NOTE: the PE branch captures effort sampling
#          uncertainty but NOT CPUE sampling uncertainty, so a PE sub-season's
#          interval is a lower bound; the BSS posterior is the fuller
#          characterization. Branch selection is driven by bss_use_pe_for(), so if a
#          component's gate verdict changes in a future run this block follows it
#          automatically.
#   Commercial/Charter (census): point distributed by sampled-harvest share with
#          a catch-per-vessel sampling CV as the relative SE (small, high coverage).
# Fix 2 (v7.0): the PE monthly effort share is now population-aware (see the PE
# branch below). The boat uses the flat 24 h gear-hours run_pe_pooled() uses; shore keeps
# crabber-hours * day_length. This removes the prior day_length over-weighting of
# long-day summer months in the boat monthly distribution (totals were, and remain,
# unaffected; only the across-month split changed). The same formula lives in 7.8;
# the two PE-share blocks are still duplicated (with a minor na.rm difference) and
# are a candidate for consolidation into one helper in a future pass.
cat("\n", strrep("=",60), "\n MONTHLY ESTIMATES BY MODE\n", strrep("=",60), "\n")

modes_lvl <- c("Shore", "Private Boat", "Commercial/Charter")
mk_mat <- function() matrix(0, nrow = n_draws_max, ncol = length(all_months),
                            dimnames = list(NULL, all_months))
bymode_C <- setNames(lapply(modes_lvl, function(x) mk_mat()), modes_lvl)

# lognormal scale, mean 1, target relative SE (mean-preserving)
rln_scale <- function(n, rel_se) {
  rel_se <- max(rel_se, 1e-4)
  sdlog  <- sqrt(log(1 + rel_se^2))
  exp(stats::rnorm(n, -sdlog^2 / 2, sdlog))
}

for(cg in bss_catch_groups) {
  for(pop in c("shore","private_boat")) {
    mode_nm <- if(pop == "shore") "Shore" else "Private Boat"
    for(ss in subseasons) {
      label    <- paste0(pop, "_", ss$name, "_", cg)
      pe_label <- paste0(pop, "_", ss$name)
      b <- bss_all[[label]]

      days_ss <- prep_days_crab(ss$start, ss$end, params, L_eff_model = L_eff_model)
      days_ss$month_label <- format(days_ss$event_date, "%Y-%m")

      if(bss_use_pe_for(b)) {   # B1.8: PE if insufficient data OR failed scale-aware gate
        summ_ss <- prep_population_summary(dwg, pop, ss$start, ss$end, params)
        # Fix 2 (v7.0), POOL-3 (v7.6): population-aware daily effort, identical
        # rationale to 7.8. Boat uses the gear-DEPLOYMENT scale
        # (mean_count * gear_per_group * tau_boat); shore keeps crabber-hours
        # (mean_count * crabbers_per_gear * day_length). Share is normalized so
        # totals are unaffected; only the monthly split changes.
        # T4.4: shared monthly effort-share helper (03_R_functions/pe_monthly_effort_share.R).
        monthly_pe <- pe_monthly_effort_share(pop, summ_ss, days_ss, params)

        pe_catch_total <- pe_all[[pe_label]][[cg]] %||% 0
        eff_total <- pe_all[[pe_label]]$effort_total %||% NA_real_
        eff_se    <- pe_all[[pe_label]]$effort_se %||% NA_real_
        rel_se <- if(!is.na(eff_total) && eff_total > 0 && !is.na(eff_se) && eff_se > 0)
                    eff_se / eff_total else 0.30
        scale_draws <- rln_scale(n_draws_max, rel_se)   # common effort-level uncertainty

        for(m in monthly_pe$month_label) {
          if(m %in% all_months) {
            sh <- monthly_pe$share[monthly_pe$month_label == m]
            bymode_C[[mode_nm]][, m] <- bymode_C[[mode_nm]][, m] + (pe_catch_total * sh) * scale_draws
          }
        }
      } else {
        C_exp_bss   <- rstan::extract(b$fit, "C_expected")$C_expected
        n_fit_draws <- dim(C_exp_bss)[1]
        C_daily_mat <- C_exp_bss[,1,,1]
        for(m in unique(days_ss$month_label)) {
          day_idx <- which(days_ss$month_label == m)
          if(length(day_idx) == 0 || !(m %in% all_months)) next
          month_C <- if(length(day_idx)==1) C_daily_mat[, day_idx] else rowSums(C_daily_mat[, day_idx])
          idx <- rep_len(1:n_fit_draws, n_draws_max)
          bymode_C[[mode_nm]][, m] <- bymode_C[[mode_nm]][, m] + month_C[idx]
        }
      }
    }
  }
}

# Commercial/charter (census): distribute by sampled-harvest share, with a
# catch-per-vessel sampling CV as the relative SE (robust fallback if unavailable).
if(!is.null(pe_all$comm_charter$daily_est)) {
  comm_daily    <- pe_all$comm_charter$daily_est |> mutate(month_label = format(date, "%Y-%m"))
  comm_total    <- pe_all$comm_charter$Dungeness_Kept %||% 0
  total_sampled <- sum(comm_daily$est_dung)
  rel_se_comm <- tryCatch({
    cpv <- dwg$interview |> filter(population == "comm_charter") |> pull(dungeness_kept)
    cpv <- cpv[is.finite(cpv)]
    if(length(cpv) > 1 && mean(cpv) > 0) (sd(cpv) / sqrt(length(cpv))) / mean(cpv) else 0.12
  }, error = function(e) 0.12)
  scale_comm <- rln_scale(n_draws_max, rel_se_comm)
  if(total_sampled > 0) {
    for(m in unique(comm_daily$month_label)) {
      if(m %in% all_months) {
        sh <- sum(comm_daily$est_dung[comm_daily$month_label == m]) / total_sampled
        bymode_C[["Commercial/Charter"]][, m] <- bymode_C[["Commercial/Charter"]][, m] + (comm_total * sh) * scale_comm
      }
    }
  }
}

monthly_bymode <- purrr::map_dfr(modes_lvl, function(md) {
  M <- bymode_C[[md]]
  tibble(
    Mode   = md,
    Month  = all_months,
    median = apply(M, 2, median),
    lo95   = apply(M, 2, quantile, probs = 0.025),
    hi95   = apply(M, 2, quantile, probs = 0.975)
  )
}) |>
  mutate(Mode  = factor(Mode, levels = modes_lvl),
         Month = factor(Month, levels = all_months),
         across(c(median, lo95, hi95), round))

print(monthly_bymode, n = nrow(monthly_bymode))
write.csv(monthly_bymode, file.path(output_dir, "monthly_estimates_by_mode.csv"), row.names = FALSE)

mode_cols <- c("Shore" = "#2c7fb8", "Private Boat" = "#d95f0e", "Commercial/Charter" = "#31a354")

# (1) Grouped (dodged) bars, one colour per mode, with 95% interval error bars.
p_bymode_dodge <- monthly_bymode |>
  ggplot(aes(x = Month, y = median, fill = Mode)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75, alpha = 0.85) +
  geom_errorbar(aes(ymin = lo95, ymax = hi95),
                position = position_dodge(width = 0.8), width = 0.3) +
  scale_fill_manual(values = mode_cols) +
  labs(title = "Monthly Dungeness Catch by Crabbing Mode",
       subtitle = "Median with 95% interval. Shore: BSS posterior. Private Boat: BSS posterior (all-gear), PE point with design-based effort SE (pot closure). Commercial/Charter: census, catch-per-vessel CV.",
       x = "Month", y = "Dungeness crab caught", fill = "Mode") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "top")
ggsave(file.path(output_dir, "plot_monthly_catch_by_mode.png"), p_bymode_dodge, width = 13, height = 6)

# (2) Faceted companion (free y) so shore and commercial panels stay readable
# next to the larger boat panel.
p_bymode_facet <- monthly_bymode |>
  ggplot(aes(x = Month, y = median, fill = Mode)) +
  geom_col(width = 0.7, alpha = 0.85) +
  geom_errorbar(aes(ymin = lo95, ymax = hi95), width = 0.3) +
  scale_fill_manual(values = mode_cols) +
  facet_wrap(~Mode, ncol = 1, scales = "free_y") +
  labs(title = "Monthly Dungeness Catch by Crabbing Mode",
       subtitle = "Median with 95% interval (free y-axis per panel)",
       x = "Month", y = "Dungeness crab caught") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")
ggsave(file.path(output_dir, "plot_monthly_catch_by_mode_facet.png"), p_bymode_facet, width = 12, height = 9)


# ===========================================================================
```

```{r monthly-mode-show, echo=TRUE, fig.height=7}
for (nm in c('p_bymode_dodge','p_bymode_facet')) if (exists(nm)) print(get(nm))
```

## 9. Catch by crabbing mode

Total season catch attributed to each crabbing mode.

```{r catch-mode, results='hide'}
# 7.9 CATCH BY CRABBING MODE
# ===========================================================================

cat("\n", strrep("=",60), "\n CATCH BY CRABBING MODE\n", strrep("=",60), "\n")

mode_table <- tibble(
  Mode = c("Shore (Dock/Jetty/Beach)", "Private Boat", "Commercial/Charter", "PORT TOTAL"),
  PE_Catch = c(
    pe_pop_sum("shore", "Dungeness_Kept"),
    pe_pop_sum("private_boat", "Dungeness_Kept"),
    pe_field("comm_charter", "Dungeness_Kept"), NA
  ),
  PE_Effort = c(
    pe_pop_sum("shore", "effort_total"),
    pe_pop_sum("private_boat", "effort_total"),
    pe_field("comm_charter", "effort_total"), NA
  )
)
mode_table$PE_Catch[4] <- sum(mode_table$PE_Catch[1:3], na.rm = TRUE)
mode_table$PE_Effort[4] <- sum(mode_table$PE_Effort[1:3], na.rm = TRUE)
print(mode_table |> mutate(across(where(is.numeric), round)))
write.csv(mode_table, file.path(output_dir, "catch_by_mode.csv"), row.names = FALSE)


# ===========================================================================
```

```{r catch-mode-show, echo=TRUE}
report_table(mode_table, 'Season catch by crabbing mode.')
```

### 9.1 Catch by gear type

In the pooled model, gear-type catch is derived after estimation by allocating each population's total over gear types using the interview gear-catch shares. As of the 2026-07-13 change (item 3 / T2.2) this allocation propagates uncertainty from both sources: the gear shares are drawn from a Dirichlet posterior (`pi ~ Dirichlet(observed gear catch + 0.5)`) and multiplied by the population total's BSS posterior draws, so each gear carries a 95% credible interval rather than a point. It is still an allocation of the pooled total, not an independently modeled per-gear CPUE process (that is the gear-resolved model's role), but its uncertainty is now honest.

```{r catch-gear, results='hide', fig.show='hide'}
# 7.10 CATCH BY GEAR TYPE (Dirichlet pi_gear + BSS total propagation; item 3)
#
# Pooled model: uses interview data to estimate the proportion of catch
# by primary gear type, then applies to combined catch estimates.
# ===========================================================================

cat("\n", strrep("=",60), "\n CATCH BY GEAR TYPE (proportional)\n", strrep("=",60), "\n")

gear_catch <- dwg$interview |>
  filter(population %in% c("shore","private_boat"), dungeness_kept > 0) |>
  mutate(
    gear_primary = case_when(
      str_detect(gear_type, "(?i)^pot|, pot|pot,") & !str_detect(gear_type, "(?i)slip ring") ~ "Pot",
      str_detect(gear_type, "(?i)ring net")     ~ "Ring Net",
      str_detect(gear_type, "(?i)trap|star")    ~ "Trap",
      str_detect(gear_type, "(?i)snare")        ~ "Snare",
      TRUE                                      ~ "Other"
    )
  )

gear_props <- gear_catch |>
  group_by(population, gear_primary) |>
  summarise(total_catch = sum(dungeness_kept), n_int = n(), .groups = "drop") |>
  group_by(population) |>
  mutate(prop_catch = total_catch / sum(total_catch)) |>
  ungroup()

cat("\n  Observed catch proportions by gear type:\n")
print(gear_props |> select(population, gear_primary, n_int, total_catch, prop_catch) |>
        mutate(prop_catch = round(prop_catch, 3)))

total_shore_catch <- pe_pop_sum("shore", "Dungeness_Kept")
total_boat_catch <- pe_pop_sum("private_boat", "Dungeness_Kept")

# --- item 3 (T2.2 / B5, 2026-07-13): propagate pi_gear + total uncertainty -------------
# The old breakdown multiplied POINT interview gear-proportions by a POINT total (a double
# point estimate with no interval). It now propagates BOTH sources: (a) each population's
# gear shares as pi ~ Dirichlet(observed gear catch + 0.5), the multinomial-Dirichlet
# posterior for catch shares; (b) the population total as its BSS posterior draws (the
# summed C_expected draws of that population's fitted sub-seasons, plus the PE point for
# any PE-fallback sub-season). Per-gear catch = total_draw * pi_draw, summed over the two
# populations and the comm/charter census (all Pot), reported as median + 95% CI. NOTE:
# the gear-resolved model has no true Dirichlet either (its pi_gear scaffolding never
# reaches its Stan), so this R-side propagation is the correct fix for both tracks and
# could be back-ported. tryCatch falls back to the old point allocation if a draw object
# is unavailable, so the section can never break the knit.
gear_levels <- c("Pot", "Ring Net", "Trap", "Snare", "Other")

gear_pop_total_draws <- function(pop, n_target) {
  labs <- paste0(pop, "_", vapply(subseasons, function(z) z$name, ""), "_Dungeness_Kept")
  draws <- NULL; pe_add <- 0
  for (lab in labs) {
    b <- bss_all[[lab]]
    if (!is.null(b) && !isTRUE(b$pe_fallback) && !is.null(b$C_exp_draws)) {
      d <- as.numeric(b$C_exp_draws)
      if (is.null(draws)) draws <- d
      else { k <- min(length(draws), length(d)); draws <- draws[seq_len(k)] + d[seq_len(k)] }
    } else {
      pe_add <- pe_add + (pe_all[[sub("_Dungeness_Kept$", "", lab)]]$Dungeness_Kept %||% 0)
    }
  }
  if (is.null(draws)) return(rep(pe_add, n_target))
  draws <- draws + pe_add
  if (length(draws) >= n_target) draws[seq_len(n_target)] else sample(draws, n_target, replace = TRUE)
}

gear_summary <- tryCatch({
  set.seed(params$bss_seed)
  n_gd <- 4000
  rdir <- function(n, alpha) {  # n x length(alpha) Dirichlet draws via gamma normalization
    g <- matrix(rgamma(n * length(alpha), shape = rep(alpha, each = n), rate = 1), nrow = n)
    g / rowSums(g)
  }
  per_gear <- matrix(0, n_gd, length(gear_levels), dimnames = list(NULL, gear_levels))
  for (pop in c("shore", "private_boat")) {
    gc <- gear_props |> dplyr::filter(population == pop)
    counts <- setNames(rep(0, length(gear_levels)), gear_levels)
    if (nrow(gc) > 0) counts[as.character(gc$gear_primary)] <- gc$total_catch
    per_gear <- per_gear + rdir(n_gd, counts + 0.5) * gear_pop_total_draws(pop, n_gd)
  }
  per_gear[, "Pot"] <- per_gear[, "Pot"] + (pe_all$comm_charter$Dungeness_Kept %||% 0)
  tibble(
    gear_primary = gear_levels,
    PE_Catch = round(apply(per_gear, 2, median)),
    lo95     = round(apply(per_gear, 2, quantile, 0.025)),
    hi95     = round(apply(per_gear, 2, quantile, 0.975))
  ) |>
    dplyr::mutate(Proportion = round(PE_Catch / sum(PE_Catch), 3)) |>
    dplyr::arrange(dplyr::desc(PE_Catch))
}, error = function(e) {
  cat(sprintf("  Gear Dirichlet propagation failed (%s); using point allocation.\n", conditionMessage(e)))
  ge <- gear_props |>
    mutate(pop_total = dplyr::case_when(population == "shore" ~ total_shore_catch,
                                        population == "private_boat" ~ total_boat_catch),
           est_catch = round(pop_total * prop_catch)) |>
    dplyr::select(gear_primary, est_catch)
  ge <- bind_rows(ge, tibble(gear_primary = "Pot",
                             est_catch = round(pe_all$comm_charter$Dungeness_Kept %||% 0)))
  ge |> group_by(gear_primary) |>
    summarise(PE_Catch = sum(est_catch), .groups = "drop") |>
    mutate(lo95 = NA_real_, hi95 = NA_real_, Proportion = round(PE_Catch / sum(PE_Catch), 3)) |>
    arrange(desc(PE_Catch))
})

cat("\n--- Catch by Gear Type (Dirichlet pi_gear + BSS-total propagation) ---\n")
print(gear_summary)
write.csv(gear_summary, file.path(output_dir, "catch_by_gear_type.csv"), row.names = FALSE)

p_gear <- gear_summary |>
  mutate(gear_primary = fct_reorder(gear_primary, PE_Catch)) |>
  ggplot(aes(x = gear_primary, y = PE_Catch, fill = gear_primary)) +
  geom_col() +
  geom_errorbar(aes(ymin = lo95, ymax = hi95), width = 0.2, na.rm = TRUE) +
  geom_text(aes(label = sprintf("%s (%.0f%%)", format(PE_Catch, big.mark=","), Proportion*100)),
            hjust = -0.1, size = 3) +
  labs(title = "Estimated Dungeness Catch by Gear Type (Dirichlet pi_gear + BSS total)",
       x = "Gear Type", y = "Estimated Catch (median, 95% CI)") +
  theme_bw() + theme(legend.position = "none") + coord_flip()
ggsave(file.path(output_dir, "plot_catch_by_gear_type.png"), p_gear, width = 10, height = 6)


```

```{r catch-gear-show, echo=TRUE, fig.height=6, fig.cap='Catch by gear type.'}
report_table(gear_summary, 'Season catch by gear type: Dirichlet-propagated interview gear-shares times BSS population-total draws (median, 95% CI). An NA interval marks the point-allocation fallback.'); if (exists('p_gear')) print(p_gear)
```

## 10. Convergence report

The per-fit diagnostic table behind the gate decision: R-hat and effective sample size for the summed effort and catch, the divergent fraction, and the pass/fail flag. **Healthy:** `pass_convergence` TRUE with R-hat near 1.00 and effective sample size well above 400 on every fit that had sufficient data.

```{r convergence-report, results='hide'}
# ---- 7.11 CONVERGENCE REPORT ----
# v7.0 (B1.8): the pass/fail decision and every diagnostic column are computed once
# in 7.6b (bss_compute_gate) and stored as bss_all[[label]]$gate_info, so this section
# is now pure assembly. This guarantees the report agrees with the method actually
# used in the totals (7.7/7.8/7.8b consume the same use_bss flag). All gate
# warnings were already emitted in 7.6b at fit time. Columns now include the
# scale-aware impact (impact_C_sd, impact_E_sd; the gating criterion) and retain
# the legacy level-distortion (distortion_C, distortion_E) as reported-only.
cat("\n", strrep("=",60), "\n CONVERGENCE REPORT\n", strrep("=",60), "\n")

convergence_df <- bind_rows(lapply(names(bss_all), function(label) bss_all[[label]]$gate_info))
print(convergence_df)
write.csv(convergence_df, file.path(output_dir, "convergence_report.csv"), row.names = FALSE)


```

```{r convergence-report-show, echo=TRUE}
# Curated, HTML-friendly view of the convergence report: short fit labels and the key
# gate columns only. The full ~20-column table (treedepth, level-distortion, and the
# per-criterion pass flags) is written to convergence_report.csv.
conv_disp <- tryCatch(
  dplyr::transmute(convergence_df,
    Fit             = short_fit(fit),
    AR              = ar_resolution,
    Div             = divergences,
    `Div frac`      = round(divergence_fraction, 3),
    `R-hat C`       = round(C_sum_rhat, 3),
    `R-hat E`       = round(E_sum_rhat, 3),
    `min n_eff`     = pmin(C_sum_neff, E_sum_neff),
    `Impact C (SD)` = round(impact_C_sd, 3),
    `Impact E (SD)` = round(impact_E_sd, 3),
    Pass            = pass_convergence,
    Method          = method_selected),
  error = function(e) convergence_df)
report_table(conv_disp,
  paste("Per-fit convergence summary (key gate columns; full detail in",
        "convergence_report.csv). Div frac is the divergent-transition fraction (backstop",
        "0.05); Impact C/E (SD) is the scale-aware gating criterion (threshold 0.10);",
        "Method is what the totals actually use. Healthy: R-hat near 1.00, min n_eff",
        "above 400, Pass = TRUE."),
  digits = 3)
```

### 10.1 Per-fit BSS plots

Per-component season-total posterior densities (C_sum, E_sum), the BSS daily effort ribbon, and the BSS daily expected-catch ribbon, one set per fitted component. Each is also saved as a PNG in the run folder.

```{r bss-fit-plots-show, echo=FALSE, fig.cap="Per-fit BSS effort, catch, and posterior-density plots (one set per fitted component)."}
# Embed the per-fit BSS plots the fit loop ggsave'd. The fit loop runs fig.show='hide'
# and overwrites each plot object every iteration, so the objects do not survive the loop
# for on-page printing; the saved PNGs do. This surfaces them (posteriors, effort, catch)
# in the HTML, matching the gear-resolved report.
bss_fit_pngs <- list.files(output_dir,
  pattern = "^plot_bss_(effort|catch|posteriors)_.*\\.png$", full.names = TRUE)
if (length(bss_fit_pngs)) knitr::include_graphics(sort(bss_fit_pngs)) else
  cat("_No per-fit BSS plots found (no fitted components this run)._")
```

## 11. Model diagnostics

Per-fit diagnostics are written to the run folder: structural-parameter summaries, divergence localization, and posterior-predictive calibration (PPC/PIT). **Healthy:** PPC coverage near nominal (about 0.95 at the 95 percent level) and PIT means near 0.5. These files are the deeper check behind the gate.

```{r diag-perfit, results='hide'}
# ---- 7.12 PER-FIT MODEL DIAGNOSTICS (B1.6) ----
# Additive per-fit diagnostics: structural parameters, divergence localization,
# and posterior predictive calibration. write_bss_diagnostics() is sourced from
# 03_R_functions/model_diagnostics.R and is fully tryCatch-wrapped, so a diagnostic
# failure prints a note and never interrupts the run. Storing bss_data with each
# fit (Section 7) is what enables the PPC; without it the PPC is skipped cleanly.
cat("\n", strrep("=",60), "\n MODEL DIAGNOSTICS\n", strrep("=",60), "\n")
for(label in names(bss_all)) {
  b <- bss_all[[label]]
  if(is.null(b$fit)) next                  # PE-only entry: nothing to diagnose
  cat(sprintf("  %s:\n", label))
  write_bss_diagnostics(b$fit,
                        if(!is.null(b$bss_data)) b$bss_data else NULL,
                        label, output_dir)
}


```

```{r diag-perfit-show, echo=FALSE}
# Surface the posterior-predictive calibration written above so the health check is on
# the page, not only in ppc_calibration_<fit>.csv. Healthy: 95% coverage near 0.95 and
# PIT mean near 0.50; a coverage far from nominal or a PIT far from 0.50 flags a
# mis-calibrated (usually over-dispersed) predictive for that data stream.
ppc <- read_fit_outputs("ppc_calibration")
ppc_disp <- if (!is.null(ppc)) tryCatch(
  dplyr::transmute(ppc,
    Fit         = fit_lbl,
    Stream      = data_type,
    `Cover 50%` = round(coverage_50, 2),
    `Cover 95%` = round(coverage_95, 2),
    `PIT mean`  = round(pit_mean, 2),
    `PIT sd`    = round(pit_sd, 2),
    n           = n_obs),
  error = function(e) NULL) else NULL
report_table(ppc_disp,
  paste("Posterior-predictive calibration by fit and data stream. Healthy: Cover 95%",
        "near 0.95 and PIT mean near 0.50; coverage well above nominal means the",
        "predictive is too wide (over-dispersed), the effort streams are the known case."))
```

### 11.1 Effort over-dispersion decomposition

Splits each effort-count predictive variance into its Poisson floor, negative-binomial observation over-dispersion, and latent-process parts, written per fit. It identifies which lever drives any excess effort variance before a prior or model change is considered.

```{r diag-overdisp, results='hide'}
# ---- 7.13 EFFORT OVER-DISPERSION DECOMPOSITION (T1.5) ----
# Additive per-fit diagnostic. The v7.0 PPC showed the negative-binomial
# effort model is over-dispersed (gear/trailer PIT central humps, coverage_50
# ~ 0.63-0.75 vs the nominal 0.50); that says the effort predictive is too
# wide but not which part of the model to tune. This decomposes each effort-
# count predictive variance, by the law of total variance, into its Poisson
# floor E[mu], the NB observation over-dispersion E[mu^2/r_E] (the r_E /
# sigma_r_E lever), and the latent process variance Var(mu) (the sigma_eps_E /
# single-cell level lever), and reports their shares so the lever is identified
# before any prior or model change. write_effort_overdispersion_diag() is
# sourced from 03_R_functions/diagnose_effort_overdispersion.R and is fully
# tryCatch-wrapped; it writes only CSVs (effort_overdispersion_decomp_<label>
# and ..._byobs_<label>) and changes nothing in the model. It needs the in-
# memory fit and its bss_data (stored with each fit in Section 7), which is why
# it runs here in the fit session rather than from the saved summary CSVs.
cat("\n", strrep("=",60), "\n EFFORT OVER-DISPERSION DECOMPOSITION (T1.5)\n", strrep("=",60), "\n")
for(label in names(bss_all)) {
  b <- bss_all[[label]]
  if(is.null(b$fit)) next                  # PE-only entry: nothing to decompose
  write_effort_overdispersion_diag(b$fit,
                                   if(!is.null(b$bss_data)) b$bss_data else NULL,
                                   label, output_dir)
}


```

```{r diag-overdisp-show, echo=FALSE}
# Surface the effort over-dispersion decomposition written above. Shares sum to 1 and
# name the lever for any excess effort variance: the Poisson floor, NB observation
# over-dispersion (r_E / sigma_r_E), or latent AR-process variance (sigma_eps_E).
od <- read_fit_outputs("effort_overdispersion_decomp")
od_disp <- if (!is.null(od)) tryCatch(
  dplyr::transmute(od,
    Fit           = fit_lbl,
    Stream        = data_type,
    Poisson       = round(share_poisson, 2),
    `NB overdisp` = round(share_nb_overdisp, 2),
    `Latent AR`   = round(share_latent, 2),
    `Cover 50%`   = round(ppc_coverage_50, 2)),
  error = function(e) NULL) else NULL
report_table(od_disp,
  paste("Effort-count predictive variance decomposition (shares sum to 1): Poisson floor,",
        "negative-binomial observation over-dispersion, and latent AR-process variance.",
        "The dominant share is the lever to tune if the effort predictive is too wide."))
```

### 11.2 Extended per-fit outputs and PSIS-LOO

The extended output series (parameter posteriors, AR paths, coverage maps, PPC residuals, sampler diagnostics, prior-vs-posterior tables) and PSIS-LOO cross-validation are written to the run folder for later analysis; nothing here changes an estimate.

```{r diag-extended, results='hide'}
# ---- 7.13b EXTENDED RUN OUTPUTS (O1-O11) ----
# Additive persistence of in-memory quantities not previously saved, so the R
# workspace is rarely needed for later diagnosis. Per-fit (O1-O6, O8, O9): full
# parameter posterior summary, AR latent path, AR-period data coverage, modeled
# daily CPUE, per-observation PPC residuals (exact randomized PIT), HMC sampler
# diagnostics (E-BFMI), summed-quantity draws, prior-vs-posterior. Run-level (O7,
# O10, O11): monthly PE-vs-BSS by mode, gear proportions, per-fit data summary.
# Sourced from 03_R_functions/save_run_diagnostics.R; every writer is tryCatch-
# wrapped and writes only CSVs, so this changes no estimate. The fit objects
# (O12) are deliberately not saved, to avoid output/ bloat.
cat("\n", strrep("=",60), "\n EXTENDED RUN OUTPUTS (O1-O11)\n", strrep("=",60), "\n")
for(label in names(bss_all)) {
  b <- bss_all[[label]]
  if(is.null(b$fit)) next                  # PE-only entry: nothing to extract
  write_fit_extended_diagnostics(b$fit, b$bss_data, b$days_ss, label, output_dir)
  write_loo_diagnostics(b$fit, b$bss_data, b$days_ss, label, output_dir)
  # POOL-5: CPUE estimator triad, saturation exponent, effort-linearity, and the
  # E/h unit assertion (03_R_functions/bss_cpue_diagnostics.R). Diagnostic-only:
  # writes cpue_estimators_/cpue_saturation_/cpue_linearity_<label>.csv and changes
  # no estimate. This is the check that surfaces a boat/shore effort-unit defect
  # (e.g. catch flat in soak time) the same way it did for gear-resolved. Wrapped
  # so a diagnostic error (or the unit assertion) can never abort the run.
  tryCatch(write_cpue_diagnostics(b, label, output_dir),
           error = function(e)
             cat(sprintf("  CPUE diagnostics skipped for %s: %s\n", label, conditionMessage(e))))
}
write_run_level_diagnostics(bss_all, pe_all,
                            if(exists("gear_props")) gear_props else NULL,
                            params, output_dir)


```

```{r diag-extended-show, echo=FALSE}
# CPUE effort-unit validity, per-fit data coverage, and PSIS-LOO, read back from the
# CSVs written above so the key diagnostics render on the page instead of only on disk.

# (a) CPUE estimator triad + linearity + saturation: is the effort unit harvest-valid?
.est <- read_fit_outputs("cpue_estimators")
.lin <- read_fit_outputs("cpue_linearity")
.sat <- read_fit_outputs("cpue_saturation")
cpue_disp <- if (!is.null(.est)) tryCatch({
  d <- dplyr::select(.est, fit_lbl, effort_unit, n_obs, cpue_ratio_of_sums, cpue_model_implied)
  if (!is.null(.lin)) d <- dplyr::left_join(d, dplyr::select(.lin, fit_lbl, beta_h, linearity_flag), by = "fit_lbl")
  if (!is.null(.sat)) d <- dplyr::left_join(d, dplyr::select(.sat, fit_lbl, sat_beta = beta, saturation_flag), by = "fit_lbl")
  dplyr::transmute(d,
    Fit          = fit_lbl,
    Unit         = effort_unit,
    n            = n_obs,
    `CPUE RoS`   = round(cpue_ratio_of_sums, 2),
    `CPUE model` = round(cpue_model_implied, 2),
    beta_h       = round(beta_h, 2),
    `linear?`    = dplyr::if_else(linearity_flag, "FLAG", "ok", missing = "-"),
    `sat beta`   = round(sat_beta, 2),
    `sat?`       = dplyr::if_else(saturation_flag, "FLAG", "ok", missing = "-"))
}, error = function(e) NULL) else NULL
report_table(cpue_disp,
  paste("CPUE effort-unit validity. beta_h is the catch-vs-effort elasticity (harvest-",
        "unbiased when it covers 1, shown as 'linear? = ok'); the model CPUE should track",
        "the design-based ratio-of-sums (RoS). 'sat? = FLAG' (catch sub-linear in soak",
        "time) is expected for pots and is the reason the unit is deployments, not hours."))

# (b) Per-fit data coverage: how much the BSS must lean on the process model.
.fds <- tryCatch(utils::read.csv(file.path(output_dir, "fit_data_summary.csv"),
                                 stringsAsFactors = FALSE), error = function(e) NULL)
fds_disp <- if (!is.null(.fds)) tryCatch(
  dplyr::transmute(.fds,
    Fit          = short_fit(fit),
    Method       = method,
    AR           = ar_resolution,
    Days         = D,
    `Eff obs`    = n_effort_obs,
    Interviews   = n_interviews,
    `I/E obs`    = n_ie_obs,
    `% days eff` = round(pct_days_with_effort, 0),
    `% days int` = round(pct_days_with_interview, 0)),
  error = function(e) NULL) else NULL
report_table(fds_disp,
  paste("Per-fit data coverage: latent days, effort observations, interviews, and I/E",
        "observations, with the percent of days carrying any effort/interview data. Thin",
        "coverage is where the BSS leans hardest on the AR process and priors."))

# (c) PSIS-LOO by stream: predictive fit and reliability (Pareto k).
.loo <- read_fit_outputs("loo_summary")
loo_disp <- if (!is.null(.loo)) tryCatch(
  dplyr::transmute(.loo,
    Fit             = fit_lbl,
    Stream          = stream,
    n               = n_obs,
    elpd_loo        = round(elpd_loo, 1),
    p_loo           = round(p_loo, 1),
    `Pareto k>0.7`  = `n_pareto_k_gt_0.7`),
  error = function(e) NULL) else NULL
report_table(loo_disp,
  paste("PSIS-LOO by data stream. p_loo is the effective number of parameters; a large",
        "'Pareto k>0.7' count means LOO is unreliable for that stream (the model is near-",
        "saturated there), so read its elpd_loo with caution."))
```

### 11.4 Summer-extrapolation transparency (T1.1)

How much of each fitted component's BSS catch falls on days with **no interview**, where the CPUE is imputed by the AR process and priors rather than observed. A high extrapolated share means that component's harvest rests on the model filling unsampled time; it is the transparency counterpart to the summer-dominance caveat, and the place to be most cautious about the headline number. The share is computed from summed daily medians, so read it as a coverage diagnostic, not a re-estimate of the total.

```{r extrapolation-transparency, echo=FALSE}
# T1.1: decompose each fit's BSS catch into interview-days vs extrapolated (no
# interview) days, from the persisted daily catch (bss_daily_catch_*) and daily CPUE
# (bss_daily_cpue_*, which carries n_interviews) outputs. A day is "extrapolated" when
# it has zero interviews, so its CPUE comes entirely from the AR process and priors.
# Diagnostic only; writes extrapolation_transparency.csv.
.dc <- read_fit_outputs("bss_daily_catch")
.dq <- read_fit_outputs("bss_daily_cpue")
extrap_tbl <- NULL
if (!is.null(.dc) && !is.null(.dq) && "n_interviews" %in% names(.dq)) {
  extrap_tbl <- tryCatch({
    dplyr::left_join(
      dplyr::transmute(.dc, fit_lbl, event_date = as.character(event_date), catch = .data[["median"]]),
      dplyr::transmute(.dq, fit_lbl, event_date = as.character(event_date), n_interviews),
      by = c("fit_lbl", "event_date")) |>
      dplyr::mutate(n_interviews = dplyr::coalesce(n_interviews, 0)) |>
      dplyr::group_by(Fit = fit_lbl) |>
      dplyr::summarise(
        `days total`         = dplyr::n(),
        `days sampled`       = sum(n_interviews > 0),
        `catch total`        = round(sum(catch, na.rm = TRUE)),
        `catch extrapolated` = round(sum(catch[n_interviews == 0], na.rm = TRUE)),
        .groups = "drop") |>
      dplyr::mutate(
        `% days sampled`       = round(100 * `days sampled` / `days total`, 0),
        `% catch extrapolated` = round(100 * `catch extrapolated` / `catch total`, 1))
  }, error = function(e) NULL)
}
if (!is.null(extrap_tbl) && nrow(extrap_tbl) > 0) {
  port_extrap_pct <- round(100 * sum(extrap_tbl$`catch extrapolated`) / sum(extrap_tbl$`catch total`), 1)
  utils::write.csv(extrap_tbl, file.path(output_dir, "extrapolation_transparency.csv"), row.names = FALSE)
  report_table(extrap_tbl,
    paste0("Share of each fitted component's BSS catch on days with no interview (CPUE ",
           "imputed by the AR process). Across the fitted components, ", port_extrap_pct,
           "% of the summed BSS catch falls on extrapolated days. A high '% catch ",
           "extrapolated' flags a component whose harvest rests on the model filling ",
           "unsampled time; shares use summed daily medians, so read as a coverage ",
           "diagnostic, not a re-estimate of the total."))
} else {
  cat("_Extrapolation transparency unavailable (daily catch / CPUE outputs not found)._")
}
```

## 12. Effort and CPUE multipliers

The fitted effort and CPUE structural quantities (day-type effects, the CPUE weekend effect B1_C, and the effort multipliers) that shape the daily series.

```{r multipliers, results='hide'}
# ---- 7.14 EFFORT & CPUE MULTIPLIERS ----
cat("\n", strrep("=",60), "\n EFFORT & CPUE MULTIPLIERS\n", strrep("=",60), "\n")

multiplier_results <- list()
for(label in names(bss_all)) {
  b <- bss_all[[label]]
  if(is.null(b$pe_fallback) || !b$pe_fallback) {
    B1_draws <- rstan::extract(b$fit, "B1")$B1
    B2_draws <- rstan::extract(b$fit, "B2")$B2
    B1C_draws <- rstan::extract(b$fit, "B1_C_out")$B1_C_out

    weekend_eff_mult <- exp(B1_draws)
    holiday_eff_mult <- exp(B1_draws + B2_draws)
    weekend_cpue_mult <- exp(B1C_draws)

    multiplier_results[[label]] <- tibble(
      fit = label,
      weekend_effort_mult = sprintf("%.2fx (%.2f-%.2f)",
        median(weekend_eff_mult), quantile(weekend_eff_mult, 0.025), quantile(weekend_eff_mult, 0.975)),
      holiday_effort_mult = sprintf("%.2fx (%.2f-%.2f)",
        median(holiday_eff_mult), quantile(holiday_eff_mult, 0.025), quantile(holiday_eff_mult, 0.975)),
      weekend_cpue_mult = sprintf("%.2fx (%.2f-%.2f)",
        median(weekend_cpue_mult), quantile(weekend_cpue_mult, 0.025), quantile(weekend_cpue_mult, 0.975)),
      B1_C_median = round(median(B1C_draws), 3),
      B1_C_lo95 = round(quantile(B1C_draws, 0.025), 3),
      B1_C_hi95 = round(quantile(B1C_draws, 0.975), 3)
    )
  }
}
if(length(multiplier_results) > 0) {
  multiplier_df <- bind_rows(multiplier_results)
  cat("\n  Multipliers relative to weekdays:\n")
  for(i in 1:nrow(multiplier_df)) {
    cat(sprintf("    %s:\n      Weekend effort: %s\n      Holiday effort: %s\n      Weekend CPUE: %s\n",
                multiplier_df$fit[i],
                multiplier_df$weekend_effort_mult[i],
                multiplier_df$holiday_effort_mult[i],
                multiplier_df$weekend_cpue_mult[i]))
  }
  write.csv(multiplier_df, file.path(output_dir, "effort_cpue_multipliers.csv"), row.names = FALSE)
}


```

```{r multipliers-show, echo=TRUE}
if (exists('multiplier_df')) report_table(multiplier_df, 'Fitted effort and CPUE multipliers.')
```

### 12.1 Expansion ratios (R_G and R_G_boat)

The gear-per-crabber ratio (shore) and gear-per-boat-group ratio (boat) that carry effort onto the gear-deployment scale. The R_G posterior below is under the model's data-driven R_G prior by default. For the T1.3 prior-sensitivity analysis the prior is overridable from `run_config.R` (`R_G_prior_mu` / `R_G_prior_sigma`); the console log records whether the prior was data-driven or an override, and this report reflects whichever prior was active for the run. A full sensitivity is a small grid of runs (R_G at 1.0, the empirical value, and 1.5) compared on the port total, not a single knit.

```{r expansion, results='hide'}
# ---- 7.16 R_G AND R_G_boat POSTERIORS ----
cat("\n", strrep("=",60), "\n EXPANSION RATIOS (R_G, R_G_boat)\n", strrep("=",60), "\n")

expansion_results <- list()
for(label in names(bss_all)) {
  b <- bss_all[[label]]
  if(is.null(b$pe_fallback) || !b$pe_fallback) {
    R_G_draws      <- rstan::extract(b$fit, "R_G")$R_G
    # POOL-1: R_T replaced by R_G_boat (gear per boat group). Both are always-present
    # parameters now; in the population where a ratio is unused it samples its prior.
    R_G_boat_draws <- rstan::extract(b$fit, "R_G_boat")$R_G_boat

    expansion_results[[label]] <- tibble(
      fit = label,
      R_G_interpretation = sprintf("%.2f gear/crabber (%.2f-%.2f)",
        median(R_G_draws), quantile(R_G_draws, 0.025), quantile(R_G_draws, 0.975)),
      R_G_boat_interpretation = sprintf("%.2f gear/group (%.2f-%.2f)",
        median(R_G_boat_draws), quantile(R_G_boat_draws, 0.025), quantile(R_G_boat_draws, 0.975))
    )
  }
}
if(length(expansion_results) > 0) {
  expansion_df <- bind_rows(expansion_results)
  for(i in 1:nrow(expansion_df)) {
    cat(sprintf("  %s: R_G=%s, R_G_boat=%s\n",
                expansion_df$fit[i], expansion_df$R_G_interpretation[i],
                expansion_df$R_G_boat_interpretation[i]))
  }
  write.csv(expansion_df, file.path(output_dir, "expansion_ratios.csv"), row.names = FALSE)
}


```

```{r expansion-show, echo=TRUE}
if (exists('expansion_df')) report_table(expansion_df, 'Expansion-ratio posteriors (R_G, R_G_boat).')
```

## 13. PE vs BSS comparison

Side by side, the design-based PE and the Bayesian BSS for each component. **How to read it:** close agreement is reassuring; a large gap on a component means the two methods disagree there, usually because that component is data-thin, and the BSS interval width tells you how well-identified it is. Since both now run on the same effort unit, a remaining gap is a genuine model disagreement, not a unit artifact.

```{r pe-vs-bss, results='hide'}
# ---- 7.17 PE VS BSS COMPARISON ----
cat("\n", strrep("=",60), "\n PE VS BSS COMPARISON\n", strrep("=",60), "\n")

comparison <- list()
for(pop in c("shore","private_boat")) {
  for(ss in subseasons) {
    label_cg <- paste0(pop, "_", ss$name, "_Dungeness_Kept")
    pe_label <- paste0(pop, "_", ss$name)
    b <- bss_all[[label_cg]]

    pe_catch <- pe_all[[pe_label]]$Dungeness_Kept %||% 0
    pe_effort <- pe_all[[pe_label]]$effort_total %||% 0

    if(!is.null(b) && (is.null(b$pe_fallback) || !b$pe_fallback)) {
      bss_catch <- b$C_expected_sum["50%"]
      bss_effort <- b$E_sum["50%"]
      catch_diff_pct <- round((bss_catch - pe_catch) / pe_catch * 100, 1)
      effort_diff_pct <- round((bss_effort - pe_effort) / pe_effort * 100, 1)
      conv <- convergence_df |> filter(fit == label_cg)
      method <- if(nrow(conv) > 0 && conv$pass_convergence[1]) "BSS" else "PE"
    } else {
      bss_catch <- NA; bss_effort <- NA
      catch_diff_pct <- NA; effort_diff_pct <- NA
      method <- "PE"
    }

    comparison[[length(comparison)+1]] <- tibble(
      component = paste0(pop, " (", ss$display_name, ")"),
      PE_effort = round(pe_effort), BSS_effort = round(bss_effort),
      effort_diff_pct = effort_diff_pct,
      PE_catch = round(pe_catch), BSS_catch = round(bss_catch),
      catch_diff_pct = catch_diff_pct,
      method_selected = method
    )
  }
}
comparison[[length(comparison)+1]] <- tibble(
  component = "comm_charter (census)",
  PE_effort = round(pe_all$comm_charter$effort_total), BSS_effort = NA, effort_diff_pct = NA,
  PE_catch = round(pe_all$comm_charter$Dungeness_Kept), BSS_catch = NA, catch_diff_pct = NA,
  method_selected = "Census"
)
comparison_df <- bind_rows(comparison)
print(comparison_df)
write.csv(comparison_df, file.path(output_dir, "pe_vs_bss_comparison.csv"), row.names = FALSE)


```

```{r pe-vs-bss-show, echo=TRUE}
report_table(comparison_df, 'Point Estimator versus Bayesian state-space estimate, by component.')
```

## 14. Season summary

Run-level metadata: the season window, sampling coverage, interview count, how many fits passed convergence, and the model configuration used.

```{r season-summary, results='hide'}
# ---- 7.18 SEASON SUMMARY ----
cat("\n", strrep("=",60), "\n SEASON SUMMARY\n", strrep("=",60), "\n")

season_days <- as.integer(as.Date(params$est_date_end) - as.Date(params$est_date_start) + 1)
total_effort_days <- n_distinct(c(dwg$shore_effort$event_date, dwg$boat_effort$event_date))
total_interview_n <- nrow(dwg$interview)
n_bss_fits <- sum(!sapply(bss_all, function(b) !is.null(b$pe_fallback) && b$pe_fallback))
n_pass <- sum(convergence_df$pass_convergence, na.rm=TRUE)

cat(sprintf("\n  Season: %s to %s (%d days)\n", params$est_date_start, params$est_date_end, season_days))
cat(sprintf("  Effort days sampled: %d (%d%%)\n", total_effort_days, round(total_effort_days/season_days*100)))
cat(sprintf("  Total interviews: %d\n", total_interview_n))
cat(sprintf("  BSS convergence: %d of %d fits passed\n", n_pass, n_bss_fits))

season_summary <- tibble(
  metric = c("Model configuration", "Season start", "Season end", "Season days",
             "Effort days sampled", "Total interviews", "BSS fits attempted", "BSS fits passed",
             "L_effective method", "AR temporal resolution"),
  value = c("Adaptive AR + L uncertainty", params$est_date_start, params$est_date_end,
            as.character(season_days), as.character(total_effort_days),
            as.character(total_interview_n), as.character(n_bss_fits), as.character(n_pass),
            if(!is.null(L_eff_model)) L_eff_model$method else "civil_twilight",
            paste(unique(na.omit(unlist(ar_resolutions))), collapse = ", "))
)
write.csv(season_summary, file.path(output_dir, "season_summary.csv"), row.names = FALSE)


```

```{r season-summary-show, echo=TRUE}
report_table(season_summary, 'Season summary and run configuration.')
```

## 15. Run parameters and timing

Writes the full run parameters and the timing summary to the run folder for reproducibility.

```{r save-params, results='hide'}
# ---- 7.19 SAVE PARAMETERS ----
timer_stop("6_outputs")
writeLines(capture.output(str(params)), file.path(output_dir, "run_parameters.txt"))

# B1.6: session and version capture for reproducibility. Pins package and Stan
# versions and the seed with each output set, so a result is reproducible from
# the committed artifacts.
writeLines(capture.output(sessionInfo()), file.path(output_dir, "session_info.txt"))
cat(sprintf("\nrstan %s; StanHeaders %s; bss_seed %s\n",
            utils::packageVersion("rstan"),
            utils::packageVersion("StanHeaders"),
            params$bss_seed),
    file = file.path(output_dir, "session_info.txt"), append = TRUE)

cat(sprintf("\nAll outputs saved to: %s\n", output_dir))
cat(paste(" ", list.files(output_dir), collapse="\n"), "\n")

# ---- TIMING SUMMARY ----
run_total <- difftime(Sys.time(), run_start_time, units = "mins")
cat("\n", strrep("=",60), "\n TIMING SUMMARY\n", strrep("=",60), "\n")
timing_df <- tibble(
  Section = names(bss_timer_log()),
  Minutes = sapply(bss_timer_log(), function(x) round(as.numeric(x$elapsed), 1))
)
print(timing_df, n = 20)
cat(sprintf("\n  TOTAL RUNTIME: %.1f minutes (%.1f hours)\n",
            as.numeric(run_total), as.numeric(run_total)/60))

# save.image(file = file.path(output_dir, "BSS-GH-pooled-CPUE-model_workspace.RData"))
```

## Appendix: divergence diagnostics (manual)

These are run by hand after a fit to localize any divergences; they are not part of the automated pipeline.

```{r divergence-diagnostics, results='hide'}

# load('output/20260619/pooled-CPUE/BSS-GH-pooled-CPUE-model_workspace.RData')

source(here::here("03_R_functions/divergence_diagnostic.R"))
# Robust to the sub-season set (build_subseasons): iterate over every fitted BSS
# component and skip PE-fallback components that have no Stan fit, rather than
# hardcoding fit labels (which omitted private_boat_ring_net_only and would miss the
# pre/post all-gear keys of a mid-season closure).
for (nm in names(bss_all)) {
  b <- bss_all[[nm]]
  if (is.null(b) || isTRUE(b$pe_fallback) || is.null(b$fit)) next
  cat(sprintf("\n=== Divergence diagnostics: %s ===\n", nm))
  diagnose_divergences(b$fit)
}
```

### 11.3 Divergence localization

Where the divergent transitions (if any) concentrate. **Healthy:** small `|SMD|` across the board, meaning divergences are diffuse and not biasing any particular parameter or total.

```{r divergence-diagnostics-show, echo=FALSE}
# Surface the top divergence-localization drivers written by the per-fit diagnostics.
# Large |SMD| (standardized median difference between divergent and bulk draws) points
# at the parameter/geometry the divergences concentrate in (e.g. a hierarchical funnel).
dl <- read_fit_outputs("divergence_localization")
dl_disp <- if (!is.null(dl) && "smd" %in% names(dl)) tryCatch(
  dl |>
    dplyr::filter(!is.na(smd)) |>
    dplyr::group_by(fit_lbl) |>
    dplyr::slice_max(abs(smd), n = 3, with_ties = FALSE) |>
    dplyr::ungroup() |>
    dplyr::transmute(Fit = fit_lbl, Parameter = param, SMD = round(smd, 2),
                     `median (bulk)` = round(median_bulk, 2),
                     `median (div)`  = round(median_div, 2)),
  error = function(e) NULL) else NULL
report_table(dl_disp,
  paste("Top divergence-localization drivers per fit (largest |SMD|). A large |SMD| is",
        "where divergent draws depart most from the bulk; it names the geometry to watch.",
        "Small values across the board mean divergences are diffuse, not biasing a total."))
```




